---
title: "Requesting GBIF data for a species list"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Requesting GBIF data for a species list}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

You have a list of species names and you want their occurrence records from
GBIF, the Global Biodiversity Information Facility. GBIF does not serve records
by name. It serves them by *taxon key*, an integer identifying one taxon in the
GBIF backbone, so the list has to be resolved to keys before anything can be
requested.

`gbif_request()` does that in one call: it matches the names against the local
GBIF backbone, collects every accepted key the matches resolve to, and hands
them to [rgbif](https://CRAN.R-project.org/package=rgbif). rgbif is needed only
for the request itself; resolving names and reading off their keys works
offline without it.

```{r}
library(taxify)

spp <- c("Quercus robur", "Pinus sylvestris", "Bellis perennis", "Morus alba")

gbif_request(spp, dry_run = TRUE)
#> 11 GBIF key(s) from 4 matched name(s). About 4,090,085 occurrence record(s) across them.
#>  [1] 2878688 7911626 7586523 5285637 7718215 8116613 9079676 8251433 7393648
#> [10] 3117424 5361889
```

Four names, eleven keys.

## Why one name gives several keys

GBIF files a name under more than one key in two situations: a homonym, where
two authors published the same name for different taxa, and a name held once as
an accepted record and again as a doubtful or duplicate one. `taxify_ids()`
lays those out, one row per name and key, with the occurrence count each key
carries:

```{r}
taxify(spp, backbone = "gbif") |>
  taxify_ids()
#>          input_name accepted_id    accepted_name taxonomic_status n_occurrences is_pick
#> 1     Quercus robur     2878688    Quercus robur         ACCEPTED       1725528    TRUE
#> 2     Quercus robur     7911626    Quercus robur         DOUBTFUL             2   FALSE
#> 3     Quercus robur     7586523    Quercus robur         DOUBTFUL             9   FALSE
#> 4  Pinus sylvestris     5285637 Pinus sylvestris         ACCEPTED       1204947    TRUE
#> 5  Pinus sylvestris     7718215 Pinus sylvestris         DOUBTFUL             0   FALSE
#> 6  Pinus sylvestris     8116613 Pinus sylvestris         DOUBTFUL             0   FALSE
#> 7  Pinus sylvestris     9079676 Pinus sylvestris         DOUBTFUL             0   FALSE
#> 8  Pinus sylvestris     8251433 Pinus sylvestris         DOUBTFUL             0   FALSE
#> 9  Pinus sylvestris     7393648 Pinus sylvestris         DOUBTFUL             0   FALSE
#> 10  Bellis perennis     3117424  Bellis perennis         ACCEPTED       1107989    TRUE
#> 11       Morus alba     5361889       Morus alba         ACCEPTED         51610    TRUE
```

`is_pick` marks the key `taxify()` reported in `accepted_id`. The rest are the
other records of the same name. Requesting only the pick would have missed the
eleven records sitting under the two doubtful *Quercus robur* keys.

`n_occurrences` is the count taken when the backbone was built. For an accepted
key it includes the records of that taxon's synonyms and descendants, which is
what a download by that key returns, so it is a good guide to the size of a
request and not an exact promise.

## Choosing what to request

Every key is requested by default. `strict = TRUE` narrows the request to
the single key `taxify()` reported:

```{r}
gbif_request(spp, strict = TRUE, dry_run = TRUE)
#> 4 GBIF key(s) from 4 matched name(s). About 4,090,074 occurrence record(s) across them.
#> [1] 2878688 5285637 3117424 5361889
```

For this list the two choices differ by 11 records out of 4.09 million, so
asking for every key costs almost nothing and picks up records that would
otherwise be dropped. The balance shifts for a list full of homonyms, which is
what `taxify_ids()` is there to show you.

`dry_run = TRUE` returns the keys and contacts nothing, so it is the way to see
what a request would cover before committing to it.

## Setting up rgbif

Everything so far ran offline against the local backbone. Sending the request
is the part that needs rgbif:

```{r}
install.packages("rgbif")
```

`method = "search"` works with nothing further. `method = "download"` is an
authenticated call and needs a GBIF account, which is free: register at
[gbif.org](https://www.gbif.org/).

rgbif reads that account from three environment variables. Its own
documentation recommends keeping them in `.Renviron` rather than passing them
as arguments, which also keeps them out of your scripts:

```{r}
usethis::edit_r_environ()
```

Add the three lines, with no quotes and no spaces around the `=`:

```
GBIF_USER=yourname
GBIF_PWD=yourpassword
GBIF_EMAIL=you@example.org
```

Save, then restart R so the file is read. rgbif also accepts the lower-case
names `gbif_user`, `gbif_pwd` and `gbif_email` in `.Rprofile` if you prefer
that; see `?rgbif::occ_download`.

To confirm GBIF accepts the account, ask it for your own past downloads.
That call is authenticated, so it only succeeds when the credentials are right:

```{r}
rgbif::occ_download_list(limit = 5)
```

If a variable is missing, `gbif_request()` stops before contacting GBIF and
names which one. If the password is wrong, GBIF answers 401.

## Sending the request

Two methods, for two situations.

`method = "search"` sends unauthenticated searches and returns the records
directly. It needs no account, and the GBIF search API limits how deep it will
page, so it suits a look at a few taxa:

```{r}
recs <- gbif_request(spp, method = "search", strict = TRUE, limit = 50)
nrow(recs)
#> [1] 200
```

The records come back as one data frame with a `taxon_key_requested` column, so
a row can be traced back to the key that asked for it.

`method = "download"` is the default. It submits an asynchronous GBIF download,
which has no record cap and is issued a DOI you can cite. It needs a GBIF
account:

```{r}
dl <- gbif_request(spp, method = "download")
rgbif::occ_download_wait(dl)
d <- rgbif::occ_download_get(dl)
recs <- rgbif::occ_download_import(d)
```

All of the keys go into one download, not one download per key: they are sent
as a single `taxonKey IN (...)` predicate, so the whole list comes back as one
dataset under one DOI. A download of four million records is prepared on GBIF's
servers and takes a while, which is why the call returns as soon as the request
is queued rather than waiting for it.

## Linking the records back to your names

The records come back keyed by GBIF taxon, not by the name you asked for, so
they have to be traced back before you can count or group by species.
`gbif_backmatch()` does that, adding `requested_key`, `input_name` and
`accepted_name`:

```{r}
recs <- gbif_request(spp, method = "search", limit = 100)
recs <- gbif_backmatch(recs, recs)

table(recs$input_name, useNA = "ifany")
```

It takes the object `gbif_request()` returned, which carries the key table as
an attribute. A `taxify_ids()` table or a `taxify()` result works too, which is
what you need after importing a download:

```{r}
matched <- taxify(spp, backbone = "gbif")

recs <- rgbif::occ_download_import(d)
recs <- gbif_backmatch(recs, matched)
```

### Why not just join on taxonKey

Because it loses records without saying so. GBIF returns the records of a key's
descendants along with its own, and a record identified below species level
carries its own key. Requesting *Pinus nigra* (5284809) returns records whose
`taxonKey` is 5686674, *P. nigra* subsp. *salzmannii*; those rows carry the
requested key in `speciesKey` and nowhere else.

In a 300-record sample of that request, 50 rows are identified to a subspecies.
Joining on `taxonKey` matches 250 of 300. `gbif_backmatch()` matches all 300,
because it tries the record's keys from the most specific outwards and takes
the first that is one of the keys you requested.

Rows matching no requested key keep `NA` in the three added columns and are
counted in a message rather than dropped.

## Citing the download

A GBIF download gets a DOI once it is ready, and citing it is what lets someone
else retrieve the same records. The download key travels with the records, so
`cite()` reports the DOI next to the backbone the names were matched against:

```{r}
recs <- gbif_backmatch(rgbif::occ_download_import(d), dl)

cite(recs)
#> ── taxify citations ────────────────────────────────────────────────
#>   [1] Colling G (2026). taxify: Offline Taxonomic Name Matching (version 0.6.0).
#>   [2] GBIF Secretariat (2024). GBIF Backbone Taxonomy. doi:10.15468/39omei
#>   [3] GBIF.org (2026-09-23) GBIF Occurrence Download https://doi.org/10.15468/dl.ckzs32
#>   ────────────────────────────────────────────────────────────
```

The DOI is read from GBIF when you call `cite()`, not stored when the request
was made, because GBIF issues it only once the download has finished
preparing. A download still running is reported as having no DOI yet.

`cite(recs, file = "refs.bib")` writes the same entries as BibTeX, the
download included.

## Starting from an already matched list

`gbif_request()` takes a `taxify()` result as well as a character vector, so
matching arguments and any filtering belong in the `taxify()` call:

```{r}
matched <- taxify(spp, backbone = "gbif", fuzzy = TRUE, kingdom = "Plantae")

matched |>
  gbif_request(method = "search", limit = 100)
```

Rows matched by another backbone are dropped with a message, because their IDs
are not GBIF keys. If nothing in the result came from GBIF, `gbif_request()`
stops and says to match against `backbone = "gbif"`.

## See also

- `taxify_ids()` for the keys and their occurrence counts on their own.

- `vignette("large-scale")` for matching lists too large to hold in memory.

- `vignette("quickstart")` for the matching itself.
