---
title: "Hybrids and aggregates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrids and aggregates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette shows how taxify matches hybrid and aggregate names and how it
attaches traits to them. Both need handling in matching and again in trait
resolution. A hybrid names a cross between two taxa, sometimes as a nothospecies
(`Mentha ×piperita`) and sometimes as a formula that spells out both parents
(`Salix alba × Salix fragilis`). An aggregate names a group of closely related
microspecies under one label (`Rubus fruticosus agg.`, `Galium mollugo s.l.`),
one step above the binomial. taxify treats the two with a parallel fallback:
where a hybrid formula has no record of its own, it resolves through its two
parents; where an aggregate has no record of its own, it falls back to the
nominal binomial.

1. **Match** the list with `taxify()`, which records hybrids in `is_hybrid` and
   `hybrid_type` and aggregates in `qualifier` and `aggregate_fallback`.
2. **Parse** the parents of a hybrid formula with `add_hybrid_info()`.
3. **Choose** whether an aggregate keeps its own concept with the `aggregates`
   argument of `taxify()`.
4. **Attach** traits with `add_trait()` or an `add_<source>()` door, which fall
   back to a hybrid's parents and an aggregate's binomial.

## Example

```{r, eval = TRUE, message = FALSE, warning = FALSE}
library(taxify)
```

### Matching a mixed species list

Botanical nomenclature marks a hybrid with the multiplication sign (×, U+00D7)
in one of three positions. A nothogenus places it before the genus name, for an
intergeneric cross (between species in two different genera), as in Leyland
cypress:

> ×Cupressocyparis leylandii

A nothospecies places it before the specific epithet, with the same genus on
both sides of the cross. Peppermint (*Mentha ×piperita*, a cross of *M.
aquatica* and *M. spicata*) is the classic case:

> Mentha ×piperita

A hybrid formula names both parent species, joined by the sign:

> Salix alba × Salix fragilis

The list below mixes an ordinary species with a nothospecies, a nothogenus, a
formula and a second nothospecies, written with the plain "x" most data use in
place of the sign, and matches them in a single call:

```{r}
names <- c(
  "Quercus robur",
  "Mentha x piperita",
  "x Cupressocyparis leylandii",
  "Salix alba x Salix fragilis",
  "Platanus x hispanica"
)

result <- taxify(names, backbone = "wfo")
result[, c("input_name", "accepted_name", "is_hybrid", "hybrid_type", "match_type")]
```

The expected output looks roughly like this:

| input_name                   | accepted_name               | is_hybrid | hybrid_type  | match_type     |
|:-----------------------------|:----------------------------|:----------|:-------------|:---------------|
| Quercus robur                | Quercus robur               | FALSE     | NA           | exact          |
| Mentha x piperita            | Mentha × piperita           | TRUE      | nothospecies | exact          |
| x Cupressocyparis leylandii  | Hesperotropsis leylandii    | TRUE      | nothogenus   | exact          |
| Salix alba x Salix fragilis  | Salix alba × Salix fragilis | TRUE      | formula      | hybrid_formula |
| Platanus x hispanica         | Platanus × hispanica        | TRUE      | nothospecies | exact          |

The two nothospecies (*Mentha*, *Platanus*) matched because WFO stores them as
accepted names with the × in the canonical name. The nothogenus ×*Cupressocyparis*
matched too: taxify tries the sign-retained forms, and WFO resolves the cross to
its accepted genus (*Hesperotropsis*). WFO carries no single record for the
*Salix* cross, so it resolves through its two parents with `match_type =
"hybrid_formula"`. Both parents resolve against WFO, so `accepted_name` holds
the accepted-parent cross "Salix alba × Salix fragilis" and `matched_name` the
input-parent cross. When a parent cannot be matched, both stay `NA`.

`is_hybrid` is `TRUE` for all four hybrid inputs, and `hybrid_type` names the
kind, whether or not the name matched: both describe the input itself and are
set during cleaning, before any backbone match.

### Hybrid parents with `add_hybrid_info()`

`hybrid_type` is already in the `taxify()` output. `add_hybrid_info()` goes
further for hybrid formulas: it parses the `input_name` column for the two
parents and resolves each against the same backbone(s) used for the result. It
adds these columns:

- `hybrid_parent_1`: the first parent binomial (for formulas) or NA
- `hybrid_parent_2`: the second parent binomial (for formulas, with
  abbreviated or omitted genera expanded) or NA
- `hybrid_parent_1_accepted`, `hybrid_parent_2_accepted`: the accepted name
  each parent resolves to, or NA if it did not match
- `hybrid_parent_1_id`, `hybrid_parent_2_id`: the backbone `taxon_id` of the
  accepted taxon each parent resolves to, ready for `id2name()`, or NA if it
  did not match

For nothogenus and nothospecies names the parent columns are NA, because the
name string does not encode the parents: those of *Mentha ×piperita* (*Mentha
aquatica* and *Mentha spicata*) appear nowhere in it. Only hybrid formulas carry
both parent names.

```{r}
result |> add_hybrid_info()
```

The new columns for the five-name example:

| input_name                   | hybrid_type   | hybrid_parent_1 | hybrid_parent_1_accepted | hybrid_parent_2 |
|:-----------------------------|:--------------|:----------------|:-------------------------|:----------------|
| Quercus robur                | NA            | NA              | NA                       | NA              |
| Mentha x piperita            | nothospecies  | NA              | NA                       | NA              |
| x Cupressocyparis leylandii  | nothogenus    | NA              | NA                       | NA              |
| Salix alba x Salix fragilis  | formula       | Salix alba      | Salix alba               | Salix fragilis  |
| Platanus x hispanica         | nothospecies  | NA              | NA                       | NA              |

### Parsing hybrid formulas

Field botanists record hybrid formulas when the parentage of a specimen is known
or suspected, so they appear in botanical and horticultural datasets. The
notation varies: some records spell out both genera in full, others abbreviate
the second genus.

```{r}
formulas <- c(
  "Salix alba x Salix fragilis",
  "Quercus pyrenaica x Q. petraea",
  "Populus nigra x Populus deltoides",
  "Rosa canina x R. gallica"
)

formula_result <- taxify(formulas, backbone = "wfo")
formula_result <- formula_result |> add_hybrid_info()

formula_result[, c("input_name", "hybrid_type",
                    "hybrid_parent_1", "hybrid_parent_2")]
```

| input_name                         | hybrid_type | hybrid_parent_1    | hybrid_parent_2     |
|:-----------------------------------|:------------|:-------------------|:--------------------|
| Salix alba x Salix fragilis        | formula     | Salix alba         | Salix fragilis      |
| Quercus pyrenaica x Q. petraea     | formula     | Quercus pyrenaica  | Quercus petraea     |
| Populus nigra x Populus deltoides  | formula     | Populus nigra      | Populus deltoides   |
| Rosa canina x R. gallica           | formula     | Rosa canina        | Rosa gallica        |

The abbreviation "Q." in the second row was expanded to "Quercus", and "R." to
"Rosa" in the fourth. The expansion is purely textual: when the second parent's
genus is a single capital letter followed by a period, it is replaced with the
first token of the first parent.

### Traits for hybrids

Trait enrichment is hybrid-aware. When you attach traits to a result that
contains a hybrid formula, each trait is resolved with a ladder:

1. the hybrid directly: if the cross resolved to a nothospecies and that name
   carries the trait, it is used as-is;
2. the average of both parents: otherwise the two parents are resolved and
   their trait values combined: numeric traits are averaged, and a categorical
   trait is taken as the shared value, or reported as `"A x B"` when the parents
   disagree (an `add_<source>()` door also warns);
3. the single available parent: if only one parent carries the trait, its value
   is used.

This happens automatically inside every `add_<source>()` door and inside
`add_trait()`; no extra call is needed.

```{r}
# A hybrid formula with no direct trait record inherits from its parents
taxify("Salix alba x Salix fragilis", backbone = "wfo") |>
  add_trait("plant_height")
```

If the parents have first been materialized with `add_hybrid_info()`, the
`add_<source>()` doors additionally expose each parent's own value as
`<trait>_parent1` / `<trait>_parent2`, so the components of the average sit
beside the combined value.

### An aggregate species list

An aggregate is marked either *agg.* (aggregate) or *s.l.* (*sensu lato*, "in
the broad sense"). Matched against WFO, which carries no aggregate taxa, each
aggregate falls back to its binomial and is flagged:

```{r}
agg_names <- c(
  "Rubus fruticosus agg.",
  "Taraxacum officinale agg.",
  "Galium mollugo s.l.",
  "Quercus robur"
)

res <- taxify(agg_names, backbone = "wfo")
res[, c("input_name", "accepted_name", "qualifier",
        "aggregate_fallback", "match_type")]
```

| input_name                | accepted_name        | qualifier | aggregate_fallback | match_type |
|:--------------------------|:---------------------|:----------|:-------------------|:-----------|
| Rubus fruticosus agg.     | Rubus fruticosus     | agg.      | TRUE               | exact      |
| Taraxacum officinale agg. | Taraxacum officinale | agg.      | TRUE               | exact      |
| Galium mollugo s.l.       | Galium mollugo       | s.l.      | TRUE               | exact      |
| Quercus robur             | Quercus robur        | NA        | NA                 | exact      |

The three aggregates resolve to their binomials with `aggregate_fallback =
TRUE`, and the plain species carries `NA` in both qualifier columns. Against
Euro+Med, where *Rubus fruticosus* aggr. is a stored taxon, the same query would
resolve to the aggregate itself with `aggregate_fallback = FALSE`.

### Traits for aggregates

Trait enrichment is aggregate-aware in the same way. The join resolves an
aggregate through the taxonomic hierarchy:

- a species query takes its own value first, and inherits its aggregate's value
  where the source records the trait only at the aggregate level;
- an aggregate query takes the aggregate-level value first, and where the source
  carries none it falls back to the nominal binomial's value, so the species'
  own measurement stands in for the group where no aggregate-level figure
  exists.

This happens automatically inside every `add_<source>()` door and inside
`add_trait()`.

```{r}
# Rubus fruticosus agg. takes the aggregate's trait where the source records
# one, otherwise the nominal Rubus fruticosus value.
taxify("Rubus fruticosus agg.") |>
  add_trait("plant_height")
```

The binomial fallback is on by default. Turning it off, per call or globally,
keeps an aggregate without aggregate-level data as `NA`:

```{r}
taxify("Rubus fruticosus agg.") |>
  add_trait("plant_height", aggregate_trait_fallback = FALSE)

options(taxify.aggregate_trait_fallback = FALSE)
```

With `options(taxify.trait_provenance = TRUE)`, each enrichment adds a
`<enrichment>_basis` column recording where each value came from: `"primary"` for
a same-level hit, `"aggregate"` for a species inheriting its aggregate's value,
and `"binomial"` for an aggregate standing in on its binomial.

## How taxify detects hybrids

Detection happens during name cleaning, before any backbone matching. Each name
is tokenized, the hybrid marker is looked for in specific positions, and the
result is classified as nothogenus, nothospecies, formula, or non-hybrid.
`is_hybrid` (logical) records whether a marker was found in the original input
and is present whether or not the name matched a backbone record; `hybrid_type`
records the finer classification (`"nothogenus"`, `"nothospecies"`, `"formula"`,
or `NA`).

In real-world data the multiplication sign is frequently replaced by a lowercase
or uppercase "x", since herbarium databases, spreadsheet exports and OCR output
rarely preserve the Unicode character. Data arrive in three forms:

1. the Unicode character itself, `×` (common in well-curated databases)
2. a lowercase `x` surrounded by spaces (common in spreadsheets and field data)
3. an uppercase `X` surrounded by spaces (less common, found in older databases
   and OCR output)

The Unicode multiplication sign is the character the International Code of
Nomenclature prescribes. taxify replaces every occurrence of it with a
space-padded "x" and works on the resulting uniform token stream, so the
downstream logic handles one representation. A standalone "x" counts as a
marker only with whitespace boundaries around it, which keeps "Saxifraga" from
triggering detection: its "x" sits inside the word.

When UTF-8 text containing × is read with a Latin-1 or Windows-1252 encoding,
the two-byte sequence can come out as "\u00c3\u0097" or "\u00c3\u2014". The name
cleaning pipeline repairs both misreadings before hybrid detection runs, so
names corrupted this way are still handled.

After detection, the marker is handled by type. For a nothospecies like "Mentha
×piperita", the cleaned form becomes "Mentha piperita", and taxify also tries
the sign-retained form "Mentha × piperita" (some backbones store the × in the
canonical name). For a nothogenus like "×Cupressocyparis leylandii", taxify
tries the sign-retained "× Cupressocyparis leylandii" and the no-space
"×Cupressocyparis leylandii" alongside the stripped binomial, so a backbone that
keeps the sign still matches.

A hybrid formula is not a single taxon name. taxify first tries the full
parent-expanded formula ("Salix alba × Salix fragilis") against the backbone:
some backbones store a cross as a name or synonym of the resulting nothospecies
(here *Salix × rubens*), and when they do, the formula resolves normally. If the
backbone does not store it, taxify parses the parents. When both parents
resolve, `matched_name` holds the input-parent formula and `accepted_name` the
accepted-parent formula; the two differ only when a parent is a synonym,
mirroring an ordinary synonym row. If either parent fails to resolve, both stay
`NA`, and the row is still flagged `is_hybrid = TRUE` with `match_type =
"hybrid_formula"`. That is as far as matching can go without a full parentage
crosswalk. The parents remain recoverable through `add_hybrid_info()`, and trait
joins fall back to them automatically.

The same-genus shorthand is recognized too: `"Salix alba × fragilis"`, with the
second parent written as a bare epithet, parses to parents "Salix alba" and
"Salix fragilis".

## What matches and what does not

The three hybrid types have different matching profiles against backbones.

Nothospecies are the best-supported form. WFO and COL both store many
nothospecies as accepted names, with the × as part of the canonical name;
*Mentha ×piperita*, *Platanus ×hispanica* and *Narcissus ×medioluteus* appear in
both. taxify tries the stripped form ("Mentha piperita") and the form with the ×
reinserted ("Mentha × piperita"), and at least one of these typically matches.

Nothogenera go through the same sign-aware logic, so coverage follows the
backbone. taxify tries "× Cupressocyparis leylandii", "×Cupressocyparis
leylandii" and the stripped "Cupressocyparis leylandii", so a backbone that
stores the sign form (as WFO and COL do) resolves the cross. When a nothogenus
is genuinely absent, the row has `match_type = "none"` and `accepted_name = NA`,
with `is_hybrid` still `TRUE`.

Hybrid formulas are tried against the backbone as a whole first. When that
misses, the row is marked `match_type = "hybrid_formula"`, the ID, rank and
classification columns stay `NA`, and the cross is named by its two parents in
`matched_name` and `accepted_name`, never collapsed to one parent. To work with
the parents, `add_hybrid_info()` resolves both against the same backbone:

```{r}
info <- result |> add_hybrid_info()
formula_rows <- info[!is.na(info$hybrid_type) & info$hybrid_type == "formula", ]
formula_rows[, c("input_name", "hybrid_parent_1", "hybrid_parent_1_accepted",
                 "hybrid_parent_2", "hybrid_parent_2_accepted")]
```

## Practical notes

WFO has the broadest coverage of plant nothospecies and nothogenera, reflecting
its focus on the world flora. COL includes hybrids across all kingdoms, with
uneven coverage. GBIF aggregates many sources and includes hybrid names where
the contributing checklists provide them. ITIS, NCBI and OTT have minimal hybrid
coverage.

Hybrid detection is input-side only. taxify detects hybrids in the names you
supply and does not scan the backbone for hybrid records. If a backbone stores
"Mentha × piperita" as an accepted name, taxify matches your input against it,
but the backbone record's own hybrid status is not exposed as a separate field;
the `is_hybrid` column reflects your input.

The formula parser expects binomials (genus plus epithet) on both sides of the ×
marker. A formula that includes a subspecies or variety rank (e.g., "Salix alba
var. vitellina × Salix fragilis") is still detected as a formula, but the parent
extraction may include the rank and infraspecific epithet in the parent name.
This is generally the desired behavior, since the full trinomial identifies the
parent more precisely than the binomial alone.

Hybrid names sometimes carry authorship strings ("Mentha ×piperita L."). The
cleaning pipeline strips authorship before matching, so an author string does
not interfere with hybrid detection or matching:

```{r}
# Authorship is stripped; hybrid detection still works
taxify("Mentha x piperita L.", backbone = "wfo")
```

`add_hybrid_info()` parses the parents from the `input_name` column and resolves
each against the backbone(s) used for the result, making one backbone lookup for
the set of unique parents (memoized within a session). The parse itself is pure
string work; the parent resolution reuses the matching engine of `taxify()`.

## How taxify matches aggregates

An aggregate groups several closely related microspecies under one name.
Apomictic complexes are the usual reason: *Rubus fruticosus*, *Taraxacum
officinale* and *Hieracium* each cover hundreds of near-indistinguishable
segregates, and field data routinely records the aggregate and leaves the
microspecies open. *agg.* is the explicit marker; *s.l.* widens a name to
include taxa a narrow reading would split off. Both mean the same thing for
matching: the name refers to the whole group. taxify folds the spelling variants
a source might use (`agg.`, `aggr.`, `agg`, `-agg`, `coll. sp.`, `sensu lato`)
to a canonical `agg.` or `s.l.` marker, recorded in the `qualifier` column. The
opposite marker, *s.str.* (*sensu stricto*, "in the narrow sense"), narrows a
name to its core binomial, so it is recorded in `qualifier` and matched as an
ordinary name.

A backbone may or may not carry a dedicated taxon for the aggregate. Only the
aggregate-bearing backbones, Euro+Med and WoRMS, store `"<binomial> aggr."` as a
concept of its own; the others record the binomial alone, without the group
above it. The `aggregates` argument sets what happens in each case.

`aggregates = "preserve"` (the default) keeps the aggregate as its own concept.
It matches the backbone's aggregate taxon where one exists and otherwise falls
back to the nominal binomial. The `aggregate_fallback` column records which
happened: `FALSE` when the dedicated aggregate taxon was found, `TRUE` when it
fell back to the binomial, and `NA` for non-aggregate names.

`aggregates = "collapse"` strips the marker up front and matches the binomial
the way any name is matched. The qualifier is still recorded, but there is no
fallback to flag, so `aggregate_fallback` stays `NA`.

## Where to go next

- [Enrichments](https://gillescolling.com/taxify/articles/enrichments.html)
  for the enrichment join and the provenance columns in general.
