---
title: "Hybrid name detection and parsing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrid name detection and parsing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Hybrid names in taxonomy

Botanical nomenclature uses a dedicated marker for hybrids: the
multiplication sign (U+00D7, rendered as \u00d7). This marker appears in
three distinct positions, each signalling a different kind of hybrid.

A **nothogenus** places the marker before the genus name, signalling an
intergeneric hybrid (a cross between species in two different genera).
Leyland cypress is a well-known example:

> \u00d7Cupressocyparis leylandii

A **nothospecies** places the marker before the specific epithet, with the
genus the same on both sides of the cross:

> Mentha \u00d7piperita

Peppermint (*Mentha \u00d7piperita*, a cross of *M. aquatica* and *M.
spicata*) is the classic case. The third form, a **hybrid formula**, names
both parent species explicitly, joined by the multiplication sign:

> Salix alba \u00d7 Salix fragilis

In real-world data, the multiplication sign is frequently replaced by a
lowercase or uppercase "x". Herbarium databases, spreadsheet exports, and
OCR outputs rarely preserve the Unicode character. taxify accepts all three
forms (`\u00d7`, `x`, `X`) and normalizes them internally. The detection
logic distinguishes a standalone "x" used as a hybrid marker from an "x"
that is part of a word (e.g., the genus *Saxifraga*) by requiring
whitespace boundaries around the letter.

```{r}
library(taxify)
```

## How taxify detects hybrids

Detection happens early in the pipeline, during name cleaning and before
any backbone matching. When `taxify()` receives an input vector, each name
passes through `clean_names()`, which calls the internal `detect_hybrid()`
function. The function tokenizes the name, looks for the hybrid marker in
specific positions, and classifies the result as nothogenus, nothospecies,
formula, or non-hybrid.

The output of `taxify()` includes an `is_hybrid` column (logical) that
records whether a hybrid marker was found in the original input. This
column is always present regardless of whether the name ultimately matched
a backbone record. The finer classification into nothogenus, nothospecies,
or formula is not exposed directly in the main output; it becomes available
through `add_hybrid_info()`, which we cover below after looking at how
matched hybrids behave in the result table.

After detection, the hybrid marker is stripped from the name before
matching. For a nothospecies like "Mentha \u00d7piperita", the cleaned
form becomes "Mentha piperita". For a hybrid formula like "Salix alba
\u00d7 Salix fragilis", only the first parent binomial ("Salix alba") is
retained as the cleaned name, since formulas are not single taxon names
and cannot match a backbone record directly.

For nothospecies, taxify also constructs a secondary search form with the
multiplication sign reinserted ("Mentha \u00d7 piperita") and attempts to
match that against the backbone. Some backbones store nothospecies with the
\u00d7 character in the canonical name, so this secondary attempt can
recover matches that the stripped form misses.

## Worked example: matching a mixed species list

Consider a list that includes ordinary species, a nothospecies, a
nothogenus, and a hybrid formula. We pass them all to `taxify()` in a
single call.

```{r}
names <- c(
  "Quercus robur",
  "Mentha x piperita",
  "x Cupressocyparis leylandii",
  "Salix alba x Salix fragilis",
  "Platanus x hispanica"
)

result <- taxify(names, backend = "wfo")
result[, c("input_name", "accepted_name", "is_hybrid", "match_type")]
```

The expected output looks roughly like this:

| input_name                   | accepted_name         | is_hybrid | match_type |
|:-----------------------------|:----------------------|:----------|:-----------|
| Quercus robur                | Quercus robur         | FALSE     | exact      |
| Mentha x piperita            | Mentha \u00d7 piperita | TRUE      | exact      |
| x Cupressocyparis leylandii  | NA                    | TRUE      | none       |
| Salix alba x Salix fragilis  | Salix alba            | TRUE      | exact      |
| Platanus x hispanica         | Platanus \u00d7 hispanica | TRUE   | exact      |

Several things are visible here. The two nothospecies (Mentha, Platanus)
matched successfully because WFO stores these as accepted names with the
\u00d7 character in the canonical name. The nothogenus
\u00d7Cupressocyparis returned no match because intergeneric hybrid genera
are less commonly included in backbone databases. The hybrid formula
matched only the first parent (Salix alba), since the formula itself is
not a single taxon name.

The `is_hybrid` column is TRUE for all four hybrid inputs, regardless of
whether the name matched. This column records a property of the input, not
of the match result.

## Extracting hybrid details with add_hybrid_info()

The `add_hybrid_info()` function takes a `taxify()` result and parses the
`input_name` column to extract structured hybrid information. It adds
three columns:

- `hybrid_parent_1`: the first parent binomial (for formulas) or NA
- `hybrid_parent_2`: the second parent binomial (for formulas, with
  abbreviated genera expanded) or NA
- `hybrid_type`: one of `"nothogenus"`, `"nothospecies"`, `"formula"`, or
  NA for non-hybrids

For nothogenus and nothospecies names, both parent columns are NA because
the input names only the hybrid itself, not its parents. The parent species
of Mentha \u00d7piperita (Mentha aquatica and Mentha spicata) are not
encoded in the name string. Only hybrid formulas carry both parent names
explicitly.

```{r}
result |> add_hybrid_info()
```

The three new columns for our five-name example:

| input_name                   | hybrid_type   | hybrid_parent_1 | hybrid_parent_2   |
|:-----------------------------|:--------------|:-----------------|:------------------|
| Quercus robur                | NA            | NA               | NA                |
| Mentha x piperita            | nothospecies  | NA               | NA                |
| x Cupressocyparis leylandii  | nothogenus    | NA               | NA                |
| Salix alba x Salix fragilis  | formula       | Salix alba       | Salix fragilis    |
| Platanus x hispanica         | nothospecies  | NA               | NA                |

## Worked example: parsing hybrid formulas

Hybrid formulas appear in botanical and horticultural datasets more often
than one might expect. Field botanists record them when the parentage of a
specimen is known or suspected. The formulas vary in notation: some spell
out both genera in full, others abbreviate the second genus.

```{r}
formulas <- c(
  "Salix alba x Salix fragilis",
  "Quercus pyrenaica x Q. petraea",
  "Populus nigra x Populus deltoides",
  "Rosa canina x R. gallica"
)

formula_result <- taxify(formulas, backend = "wfo")
formula_result <- formula_result |> add_hybrid_info()

formula_result[, c("input_name", "hybrid_type",
                    "hybrid_parent_1", "hybrid_parent_2")]
```

| input_name                         | hybrid_type | hybrid_parent_1    | hybrid_parent_2     |
|:-----------------------------------|:------------|:-------------------|:--------------------|
| Salix alba x Salix fragilis        | formula     | Salix alba         | Salix fragilis      |
| Quercus pyrenaica x Q. petraea     | formula     | Quercus pyrenaica  | Quercus petraea     |
| Populus nigra x Populus deltoides   | formula     | Populus nigra      | Populus deltoides    |
| Rosa canina x R. gallica           | formula     | Rosa canina        | Rosa gallica         |

The genus abbreviation "Q." in the second example was expanded to
"Quercus" automatically. taxify infers the full genus from the first
parent in the formula. The same expansion happened for "R." to "Rosa" in
the fourth row. This expansion is purely textual: the first token of the
first parent is used as the genus for the second parent whenever the
second parent's genus field matches the pattern of a single capital letter
followed by a period.

## What matches and what does not

The three hybrid types have different matching profiles against backbone
databases.

**Nothospecies** are the best-supported form. WFO and COL both store many
nothospecies as accepted names, with the \u00d7 character as part of the
canonical name. Mentha \u00d7piperita, Platanus \u00d7hispanica, and
Narcissus \u00d7medioluteus are examples that appear in both backbones.
taxify's matching logic handles the marker correctly: it first tries the
stripped form ("Mentha piperita") and then the form with the \u00d7
reinserted ("Mentha \u00d7 piperita"). At least one of these typically
matches.

**Nothogenera** have lower coverage. Intergeneric hybrids like
\u00d7Cupressocyparis, \u00d7Triticosecale, and \u00d7Festulolium exist in
some backbones but are absent from others. WFO includes several
nothogenera relevant to agriculture and horticulture. COL's coverage
varies by taxonomic group. When a nothogenus does not match, the output
row will have `match_type = "none"` and `accepted_name = NA`, but
`is_hybrid` will still be TRUE.

**Hybrid formulas** will not match a backbone record directly, because the
formula is not a taxon name. taxify extracts the first parent binomial as
the cleaned name for matching, so the result row reflects the match status
of the first parent. To resolve both parents, match them separately.

```{r}
# Match both parents of a hybrid formula separately
parents <- c("Salix alba", "Salix fragilis")
parent_result <- taxify(parents, backend = "wfo")
```

This approach gives a full match result (accepted name, synonym status,
authorship) for each parent individually. In a dataset with many hybrid
formulas, we can extract the parent columns from `add_hybrid_info()` and
feed them back through `taxify()` as a batch.

```{r}
# Batch-resolve all hybrid formula parents
info <- result |> add_hybrid_info()
formula_rows <- info[info$hybrid_type == "formula" & !is.na(info$hybrid_type), ]

all_parents <- unique(na.omit(c(
  formula_rows$hybrid_parent_1,
  formula_rows$hybrid_parent_2
)))

parent_matches <- taxify(all_parents, backend = "wfo")
```

## The multiplication sign and its substitutes

The Unicode multiplication sign (U+00D7) is the correct character for
hybrid notation under the International Code of Nomenclature. In practice,
data arrive with three common representations:

1. The Unicode character itself: `\u00d7` (common in well-curated databases)
2. A lowercase `x` surrounded by spaces (common in spreadsheets and field
   data)
3. An uppercase `X` surrounded by spaces (less common, but occurs in older
   databases and OCR output)

taxify normalizes all three forms internally. The `detect_hybrid()`
function replaces every occurrence of U+00D7 with a space-padded "x" and
then works with a uniform token stream, so the downstream logic only needs
to handle one representation. The space-boundary requirement prevents
false positives: "Saxifraga" does not trigger hybrid detection because the
"x" sits within a word rather than standing alone between tokens.

A subtlety arises with mojibake. When UTF-8 text containing the \u00d7
character is read with a Latin-1 or Windows-1252 encoding, the two-byte
sequence can be misinterpreted as "\u00c3\u0097" or "\u00c3\u2014". The
name cleaning pipeline detects and repairs both of these common
misreadings before hybrid detection runs, so names corrupted by encoding
errors are still handled correctly.

## Practical notes

**Which backbones have the most hybrids.** WFO has the broadest coverage
of plant nothospecies and nothogenera, reflecting its focus on the world
flora. COL includes hybrids across all kingdoms but coverage is uneven.
GBIF aggregates data from many sources and includes hybrid names where the
contributing checklists provide them. ITIS, NCBI, and OTT have minimal
hybrid coverage.

**Hybrid detection is input-side only.** taxify detects hybrids in the
names that you supply. It does not scan the backbone for hybrid records.
If a backbone stores "Mentha \u00d7 piperita" as an accepted name, taxify
will match your input against it, but the backbone record's own hybrid
status is not exposed as a separate field. The `is_hybrid` column reflects
your input, not the backbone.

**Formulas with infraspecific ranks.** The parser expects binomials (genus
plus epithet) on both sides of the \u00d7 marker. Formulas that include
subspecies or variety ranks (e.g., "Salix alba var. vitellina \u00d7 Salix
fragilis") will still be detected as formulas, but the parent extraction
may include the rank and infraspecific epithet as part of the parent name.
This is generally the desired behavior, since the full trinomial identifies
the parent more precisely than the binomial alone.

**Authorship in hybrid names.** Hybrid names sometimes carry authorship
strings (e.g., "Mentha \u00d7piperita L."). The name cleaning pipeline
strips authorship before matching, so the presence of an author string
does not interfere with hybrid detection or matching.

```{r}
# Authorship is stripped; hybrid detection still works
taxify("Mentha x piperita L.", backend = "wfo")
```

**Adding hybrid info is lightweight.** `add_hybrid_info()` operates
entirely on the `input_name` column via string parsing. It does not
re-query any backbone or access any files on disk. On a result with 10,000
rows, the function completes in milliseconds.
