---
title: "Working with large species lists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with large species lists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

taxify is designed for lists of a few hundred names and lists of a few
hundred thousand names alike. The underlying engine (vectra) stores backbone
databases in a columnar binary format (.vtr) that supports memory-mapped
access, hash-indexed lookups, and OpenMP-parallel fuzzy joins. None of this
requires special configuration from the user. But knowing how the pieces fit
together helps when tuning a workflow for speed, memory, or disk usage at
scale.

This vignette covers the performance-relevant internals, gives concrete
timing guidance for different list sizes, and walks through four worked
examples: exact vs. fuzzy matching, multi-backend fallback ordering, batch
processing of very large lists, and pre-downloading resources before a batch
run.

```{r}
library(taxify)
```


## How taxify scales

### The .vtr columnar format

Every backbone ships as a `.vtr` file: a binary columnar format written by
the vectra C11 engine. Unlike CSV or TSV, the `.vtr` format stores each
column contiguously on disk with lightweight compression. taxify never parses
text at query time. There is no `read.csv()` step, no string splitting, no
quote escaping. The backbone is already in a query-ready binary layout.

Backbones are distributed as pre-built `.vtr` files (hosted on Zenodo and
GitHub Releases), so users download a single binary file that is ready to
query immediately — no CSV parsing, no conversion step. The `.vtr` files are
typically 30-50% smaller than the original Darwin Core CSV because the
columnar layout compresses string columns more efficiently than row-oriented
text.

### Exact matching: hash-indexed lookups

When a backbone is first used in a session, vectra materializes it into an
in-memory columnar block with hash indexes on the name and genus columns.
Subsequent lookups against that block run in essentially constant time per
name.

Exact matching uses `block_lookup()`, which resolves each input name via a
hash index. This is an O(1) operation per name and the reason exact matching
scales linearly with list size. A list of 100,000 clean plant names matches
against WFO in seconds, not minutes.

The exact pipeline is more thorough than a simple string comparison. It runs
five passes in sequence, each catching a different class of name variation:

1. **Case-sensitive exact match** against the canonical name column.

2. **Case-insensitive match** against a precomputed lowercased key.

3. **Latin orthographic normalization** that maps common epithet variants
   (e.g., *-ii* to *-i*, *-anum* to *-ana*) to a canonical form.
4. **Infraspecific-to-species fallback** that strips variety/subspecies
   qualifiers and matches against the binomial.
5. **Hybrid name normalization** that resolves nothospecies formatting
   differences (e.g., *Salix x rubens* vs. *Salix xrubens*).

All five passes use hash lookups. A name that matches in pass 1 is never
tested in passes 2-5. In practice, pass 1 resolves 85-95% of names from
clean input, and passes 2-4 pick up another 2-5%. The total cost of exact
matching is dominated by the hash lookups, which are O(1) per name
regardless of backbone size.

### Fuzzy matching: genus-blocked string distance

Fuzzy matching is fundamentally more expensive. For each unmatched name,
vectra computes string distances (Damerau-Levenshtein by default) against
all backbone entries that share the same genus. This genus-blocking strategy
reduces the search space from millions of backbone entries to a few hundred
or thousand (the typical number of species per genus). The computation is
parallelized across cores via OpenMP, using 4 threads by default.

Fuzzy matching 5,000 names against WFO, where every name carries a typo, took
27 seconds in the repository's benchmark run. Large genera such as *Carex* or
*Astragalus* are more expensive per name than small ones, and the cost grows
with the number of names that fail the exact pass and with the size of the
backbone, so a larger backbone raises it proportionally.

A secondary fuzzy pass handles misspelled genera. When the genus itself is
wrong (e.g., *Qurecus* instead of *Quercus*), the genus-blocked join misses
the name entirely. taxify runs a fallback pass that blocks on the first two
characters of the name instead of the full genus. This catches most single-
character genus typos while keeping the search space much smaller than a
full cross-join.

The practical consequence: exact-only matching is fast at any scale, and
fuzzy matching is the knob that controls how long a run takes.


## Backbone loading and the session cache

The first time `taxify()` is called for a given backend, several things
happen behind the scenes. The function resolves the backbone path through a
four-step fallback: session cache, versioned directory on disk, legacy flat
directory, and finally auto-download from Zenodo if no local copy exists.
Once the path is known, vectra materializes the `.vtr` into an in-memory
columnar block and builds hash indexes on the name and genus columns. This
initialization step took about 5 seconds for WFO (1.6 million rows) in the
repository's benchmark run, and scales with backbone size from there. Every
subsequent `taxify()` call in the
same R session reuses the materialized block. There is no repeated file I/O.

Two caches operate in parallel. The path cache (`.taxify_cache`) maps
backend names to `.vtr` file paths on disk. Once a path is resolved, it
stays cached so that `ensure_backbone()` does not re-scan the file system.
The data cache (`.taxify_env`) holds the materialized columnar block itself,
keyed by file path. It also stores the session manifest, version-check
flags, enrichment paths, and coverage data for the genus register. Both are
package-level environments that persist for the duration of the R session
and are shared across all `taxify()` calls.

The first `taxify()` call in a session also triggers a version check. taxify
fetches a manifest from GitHub (a small JSON file listing the latest version
of each backbone) and compares it against the locally installed version. If
a newer backbone is available, it is downloaded automatically. This check
runs once per backend per session. Subsequent calls skip it entirely. If the
network is unavailable, the check is skipped and the local copy is used
as-is.

To see where backbones live on disk:

```{r}
taxify_data_dir()
#> [1] "C:/Users/jane/AppData/Local/R/taxify"
```

On Linux this is typically `~/.local/share/R/taxify`, on macOS
`~/Library/Application Support/R/taxify`. The path is determined by
`tools::R_user_dir("taxify", "data")` and is shared across all R projects
on the same machine. Two R sessions running on the same machine can read the
same backbone files concurrently without conflict because the `.vtr` files
are read-only at query time.


## Worked example: exact vs. fuzzy matching

The simplest performance lever is the `fuzzy` argument. When input names are
clean (e.g., names from a curated database, an existing taxonomic checklist,
or the output of a previous taxify run), disabling fuzzy matching skips the
string-distance computation entirely.

Consider a list of 10,000 plant names extracted from an herbarium database
where names are already in standard binomial form. We time both modes:

```{r}
# Assume `species_list` is a character vector of 10,000 plant names

# Exact + fuzzy (default)
t_fuzzy <- system.time(
  result_fuzzy <- taxify(species_list, backend = "wfo", fuzzy = TRUE)
)

# Exact only
t_exact <- system.time(
  result_exact <- taxify(species_list, backend = "wfo", fuzzy = FALSE)
)

t_fuzzy["elapsed"]
t_exact["elapsed"]
```

The timings depend on your list and your machine, so run the pair on your own
data rather than trusting a figure from here; `scripts/benchmark-worldflora.R`
in the taxify repository is the measured reference. The shape is consistent:
the exact-only run is roughly an order of magnitude faster. The exact pass
matches most names on the first try through its five-pass pipeline
(case-sensitive, case-insensitive, Latin orthographic normalization, infraspecific
fallback, and hybrid normalization). Fuzzy matching picks up the remaining
names with minor misspellings, but at a cost that scales with the number of
unmatched names times the average genus size in the backbone.

The ratio between the two modes depends on input quality. For a list of
names extracted from a curated database (GBIF occurrence records, a
published checklist, or a previous taxify run), the exact pass resolves
95-99% of names and the fuzzy pass adds very little. For OCR-transcribed
herbarium labels or citizen science data with frequent misspellings, the
exact pass might resolve only 70-80% and the fuzzy pass becomes essential.

A practical two-pass pattern for large lists: run exact-only first, inspect
the unmatched names, and decide whether the fuzzy pass is worth the time.

```{r}
# Pass 1: exact only
result <- taxify(species_list, backend = "wfo", fuzzy = FALSE)

# How many names remain unmatched?
n_unmatched <- sum(result$match_type == "none")
message(n_unmatched, " names unmatched after exact pass")

# Pass 2: fuzzy only on the unmatched subset
if (n_unmatched > 0) {
  unmatched_names <- result$input_name[result$match_type == "none"]
  fuzzy_result <- taxify(unmatched_names, backend = "wfo", fuzzy = TRUE)

  # Merge back
  matched_rows <- fuzzy_result$match_type != "none"
  idx <- match(fuzzy_result$input_name[matched_rows],
               result$input_name)
  result[idx, ] <- fuzzy_result[matched_rows, ]
}
```

This pattern is especially useful when only 1-5% of names need fuzzy
matching. The exact pass finishes in seconds even for 100,000 names, and the
fuzzy pass operates on a much smaller subset. The total wall time is often
less than half of what a single `fuzzy = TRUE` call would take, because the
fuzzy engine does not need to allocate working memory or build query tables
for the names that already matched exactly.

One subtlety: the second call to `taxify()` does not re-materialize the
backbone. The session cache from the first call is still active, so the
fuzzy-only pass starts immediately with the string-distance computation.
There is no penalty for splitting the work into two calls.


## Worked example: multi-backend fallback ordering

When `taxify()` receives multiple backends, it processes them as a
sequential fallback chain. Names matched by an earlier backend are excluded
from later ones. The order matters for performance: the first backend sees
all names, the second sees only those that failed, and so on.

Suppose we have a mixed-kingdom species list from a freshwater ecology
survey: mostly aquatic plants, some fish, a handful of invertebrates. WFO
covers the plants, COL covers everything but is larger and slower to search.
Putting WFO first means the plant names (the majority) are resolved quickly,
and only the animal names fall through to COL.

```{r}
# 8,000 names: ~6,000 plants, ~1,500 fish, ~500 invertebrates
t_wfo_first <- system.time(
  result_a <- taxify(survey_names,
                     backend = c("wfo", "col"),
                     fuzzy = TRUE)
)
t_wfo_first["elapsed"]

# Reversed order: COL first, WFO second
t_col_first <- system.time(
  result_b <- taxify(survey_names,
                     backend = c("col", "wfo"),
                     fuzzy = TRUE)
)
t_col_first["elapsed"]
```

Name resolution is identical either way -- both orders resolve the same names
to the same accepted names -- but WFO first is faster, because WFO's smaller
backbone (1.6 million rows) resolves most of the list before COL's larger one
(5.3 million rows) is ever touched. The saving comes from two sources: the exact pass against WFO is
faster (smaller hash table), and the fuzzy pass against COL runs on 2,000
names instead of 8,000. Since fuzzy matching cost scales linearly with the
number of unmatched names, resolving 6,000 names via WFO's fast exact pass
eliminates the need for 6,000 fuzzy comparisons against COL's much larger
genus blocks.

Note that the `backend` column in the output records which backend resolved
each name. This is useful for quality control: if a name was resolved by the
second backend in the chain, it means the first backend either did not
contain it or matched it differently. Inspecting the `backend` column after
a multi-backend run can reveal patterns in taxonomic coverage gaps.

General guidelines for backend ordering:

- Plant-only lists: `"wfo"` alone is sufficient. WFO has the most complete
  plant synonym coverage and a compact backbone.
- Marine lists: `"worms"` first, then `"col"` or `"gbif"` for anything
  WoRMS misses.
- Mixed-kingdom lists: put the backbone that covers the dominant kingdom
  first. For a list that is 80% plants and 20% animals, `c("wfo", "col")`
  is faster than `c("col", "wfo")`.
- Maximizing coverage: `c("wfo", "col", "gbif")` casts the widest net but
  involves three backbone loads. For lists under 10,000 names the extra
  loading time is negligible. For 100,000+ names, the extra fuzzy passes
  add up.


## Backbone sizes on disk

Each backbone's `.vtr` file is a one-time download stored in
`taxify_data_dir()`. The sizes below are approximate and depend on the
backbone version.

<!-- manifest:backbone-sizes -->
| Backbone | Names | Download | Version |
|---|---|---|---|
| WFO | 1.6M | 797 MB | 2026.06 |
| COL | 5.3M | 2.0 GB | 2026.06 |
| GBIF | 6.4M | 1.9 GB | 2026.06 |
| ITIS | 992k | 205 MB | 2026.06 |
| NCBI | 2.8M | 514 MB | 2026.06 |
| OTT | 3.7M | 727 MB | 3.7.3 |
| WoRMS | 1.6M | 347 MB | 2026.07 |
| Euro+Med | 147k | 35 MB | 2026.07 |
| Species Fungorum | 315k | 71 MB | 2026.06 |
| AlgaeBase | 172k | 36 MB | 2026.06 |
| FishBase | 103k | 19 MB | 2026.06 |
| SeaLifeBase | 134k | 29 MB | 2026.06 |
| Reptile Database | 50k | 10 MB | 2026.07 |
| LCVP | 1.3M | 252 MB | 3.0.1 |
| WCVP | 1.4M | 334 MB | 2026.06 |
| **All 15** | **26.1M** | **7.2 GB** | |
<!-- /manifest:backbone-sizes -->

A full installation of all fifteen backbones occupies several GB. Most
workflows need only one or two. The WFO backbone alone covers the vast
majority of plant taxonomy use cases in under a gigabyte.

The download sizes are comparable to the on-disk sizes since the `.vtr`
format is already compressed. No additional decompression step runs after
download. The file that arrives on disk is the file that vectra reads at
query time.

Enrichment files are much smaller. The largest enrichment is WCVP (native
range data, ~2 million rows) at roughly 30-40 MB. Most enrichments are
under 5 MB. A full set of <!-- manifest:enrichment-count -->90<!-- /manifest:enrichment-count --> enrichments adds a few hundred MB to disk
usage.


## Memory footprint

`scripts/benchmark-memory.R` in the taxify repository measures this, and
`scripts/benchmark-memory-results.json` records the run. It reports resident set
size rather than the R heap: vectra reads a `.vtr` through the operating system
instead of materializing it as R objects, so `gc()` sees only a fraction of what
a backbone actually costs. Each stage runs in a fresh R process, because resident
size never falls back after a peak.

Matching 5,000 names on Windows 11, R 4.6.0, taxify 0.3.21, in megabytes above an
idle R process:

| Backbone | `.vtr` | After load | Exact, 5,000 | Fuzzy, 5,000 |
|---|---|---|---|---|
| WFO | 797 | 1,443 | 1,732 | 2,760 |
| COL | 1,958 | 3,751 | 4,544 | 6,128 |
| GBIF | 1,855 | 4,024 | 4,703 | 5,910 |

Opening a backbone costs about 1.8 to 2.2 times its `.vtr` size, because the
columnar block carries hash indexes and decompressed string data alongside the
stored columns. The block persists for the session and every `taxify()` call
reuses it. A second backbone -- during a multi-backbone fallback, say -- adds its
own block; the two coexist independently.

Fuzzy matching is what dominates the transient cost. For each fuzzy pass taxify
writes a temporary `.vtr` of the unmatched names and their genera and hands it to
vectra's `fuzzy_join()`, which allocates a working buffer proportional to the
number of unmatched names times the average genus block size. On the worst case
above -- 5,000 names that all need fuzzy resolution -- that buffer reaches 1.3 GB
against WFO and 2.4 GB against COL, several times what the same names cost on the
exact path. A realistic list, where most names match exactly, pays a fraction of
it: the buffer scales with the unmatched remainder, not the input. The temporary
files and buffers are freed after each pass.

Enrichment `.vtr` files are loaded on demand and are far smaller. An enrichment
join builds a temporary `.vtr` of unique accepted names, runs an `inner_join()`
against the enrichment `.vtr`, and fills the result via `match()`. The enrichment
is never fully materialized; only the joined subset is collected, so the cost is
proportional to the number of unique accepted names in the result -- typically
much smaller than the input list, since synonyms collapse onto shared accepted
names.

If memory is tight, three strategies help:

1. Use a smaller backbone. WFO costs roughly a third of what GBIF costs, and for
   plant-only lists there is no coverage penalty.
2. Clear the cache between phases. Once matching is done and the result is saved,
   release the backbone before running downstream models.
3. Enrich inside the matching loop (the chunk-and-write pattern below) rather
   than accumulating the full result and enriching afterwards.

```{r}
# Match names
result <- taxify(species_list, backend = "gbif")

# Save result
saveRDS(result, "matched_names.rds")

# Free the backbone from memory
taxify_clear_cache()

gc()
```


## Worked example: batch processing a very large list

Lists above 100,000 names are common in biodiversity informatics. A national
herbarium digitization project might produce 500,000 label transcriptions. A
metabarcoding pipeline might output 200,000 OTU labels. Processing these in
a single `taxify()` call works, but splitting into chunks gives two
practical benefits: progress monitoring and memory stability.

taxify's matching engine handles the full vector internally, and a single
`taxify()` call on 500,000 names will produce correct results. But two
practical issues arise at this scale. First, the fuzzy-join working set
grows with input size: 500,000 names with 10% unmatched means 50,000 fuzzy
comparisons, each scanning a genus block. The temporary `.vtr` files and
distance matrices for this many comparisons can spike memory by several
hundred MB. Second, if the R process is interrupted mid-run (Ctrl+C, OOM
kill, session timeout), the entire result is lost. Chunking at 50,000-100,000
names keeps peak memory predictable and provides natural restart points.

```{r}
# 300,000 names from a herbarium digitization project
all_names <- readLines("herbarium_names.txt")
chunk_size <- 50000

# Split into chunks
chunks <- split(all_names,
                ceiling(seq_along(all_names) / chunk_size))

# Process each chunk
results <- lapply(seq_along(chunks), function(i) {
  message(sprintf("Chunk %d/%d (%d names)...",
                  i, length(chunks), length(chunks[[i]])))
  taxify(chunks[[i]], backend = "wfo", fuzzy = TRUE, verbose = FALSE)
})

# Combine
result <- do.call(rbind, results)
nrow(result)
#> [1] 300000
```

The backbone stays in memory across chunks (the session cache is not
cleared between calls), so each chunk after the first skips the
initialization overhead. Only the fuzzy-join working set is allocated and
freed per chunk.

For lists in the millions (e.g., processing all occurrence records from a
GBIF download), consider writing results to disk after each chunk rather
than accumulating in memory:

```{r}
output_dir <- "results"
dir.create(output_dir, showWarnings = FALSE)

for (i in seq_along(chunks)) {
  message(sprintf("Chunk %d/%d", i, length(chunks)))
  res <- taxify(chunks[[i]], backend = "wfo",
                fuzzy = TRUE, verbose = FALSE)
  saveRDS(res, file.path(output_dir,
                         sprintf("chunk_%04d.rds", i)))
}

# Combine when needed
all_files <- list.files(output_dir, pattern = "\\.rds$",
                        full.names = TRUE)
result <- do.call(rbind, lapply(all_files, readRDS))
```

This pattern keeps R's memory usage bounded by a single chunk regardless of
total list size. It also makes the workflow resumable: if the process dies at
chunk 47 of 60, we can check which `.rds` files exist and restart from chunk
48. Adding a simple skip condition handles this:

```{r}
for (i in seq_along(chunks)) {
  out_file <- file.path(output_dir, sprintf("chunk_%04d.rds", i))
  if (file.exists(out_file)) next
  message(sprintf("Chunk %d/%d", i, length(chunks)))
  res <- taxify(chunks[[i]], backend = "wfo",
                fuzzy = TRUE, verbose = FALSE)
  saveRDS(res, out_file)
}
```

One detail worth noting: the chunk boundaries are arbitrary and do not
affect matching quality. Each chunk is matched independently against the
backbone. A name that appears in chunk 3 gets the same result as if it
appeared in chunk 7 because the backbone is deterministic and the matching
logic is stateless across calls. The only shared state between chunks is the
session cache (the materialized backbone block), which improves performance
by avoiding repeated initialization.


## Cache management

taxify uses two internal environments for session-level caching. The path
cache stores the disk location of each loaded backbone. The data cache
stores materialized columnar blocks, the session manifest, version-check
flags, and enrichment paths. Both persist until the R session ends or the
user explicitly clears them.

`taxify_clear_cache()` removes all loaded backbone paths from memory. The
next `taxify()` call will re-read from disk and re-materialize. This is
useful after a large matching run when the backbone is no longer needed and
the memory can be reclaimed.

```{r}
# After finishing all matching work
taxify_clear_cache()
gc()
```

Clearing the cache does not delete any files from disk. The `.vtr` files
remain in `taxify_data_dir()` and will be reloaded on the next use. The
cost of reloading is the same 1-10 second initialization time that the first
call in a session incurs. For a workflow where matching is done in one
phase and downstream modelling in another, this is a worthwhile trade: spend
3 seconds reloading WFO later if needed, but free 100 MB of RAM for a
memory-intensive ordination or species distribution model.

`taxify_refresh_manifest()` is a narrower operation: it invalidates the
cached copy of the remote manifest (the JSON file listing the latest version
of each backbone and enrichment). This forces the next `taxify()` call to
re-check for updates. Normally the manifest is fetched once per session and
cached. In a long-running R session (e.g., an RStudio session that stays
open for days), calling `taxify_refresh_manifest()` before a batch run
ensures you are working against the latest backbone version. If a new
backbone release was published since the session started, the version check
will detect it and trigger an automatic download.

```{r}
taxify_refresh_manifest()
```


## Disk storage and sharing across projects

All taxify data lives under `taxify_data_dir()`, which resolves to the
platform-specific user data directory via `tools::R_user_dir("taxify",
"data")`. The layout is:

```
taxify_data_dir()/
  wfo/
    latest/
      wfo.vtr          # the backbone
      wfo.meta         # download provenance
      meta.json        # version metadata
  col/
    latest/
      col.vtr
      ...
  enrichment/
    iucn/
      latest/
        iucn.vtr
        meta.json
    zanne/
      latest/
        zanne.vtr
        meta.json
    ...
```

This directory is per-user, not per-project. A backbone downloaded once is
available to every R project on the machine without duplication. There is no
need to copy `.vtr` files into a project directory or version-control them.

If multiple users on a shared server need the same backbones, one user can
download them and the others can set the `R_USER_DATA_DIR` environment
variable (or symlink `taxify_data_dir()`) to a shared location. The `.vtr`
files are read-only at query time, so concurrent access from multiple R
sessions is safe. No file locking is needed.

To check how much disk space taxify is currently using:

```{r}
# Total size of all backbones and enrichments
data_dir <- taxify_data_dir()
files <- list.files(data_dir, recursive = TRUE, full.names = TRUE)
total_mb <- sum(file.size(files), na.rm = TRUE) / 1048576
message(sprintf("taxify data: %.0f MB across %d files",
                total_mb, length(files)))
```

To remove a specific backbone (e.g., GBIF after finishing a project that
needed it), delete its directory:

```{r}
# Remove GBIF backbone (frees ~500-700 MB)
unlink(file.path(taxify_data_dir(), "gbif"), recursive = TRUE)

# Clear the session cache so taxify() doesn't try to use the old path
taxify_clear_cache()
```

Deleting a backbone directory is safe. The next `taxify()` call for that
backend will re-download it from Zenodo if needed.


## Worked example: pre-downloading resources

For a reproducible batch pipeline (e.g., a Makefile or targets plan), it is
cleaner to separate the download step from the analysis step. Downloads can
fail due to network issues, and you want to know about that before a 2-hour
matching run starts.

`taxify_download()` downloads one or more backbone `.vtr` files.
`taxify_download_enrichment()` does the same for enrichment layers. Both
are idempotent: if the file already exists and the version is current, they
return immediately.

```{r}
# Pre-download everything needed for a multi-kingdom analysis
# with conservation status and trait enrichments

# Backbones
taxify_download(c("wfo", "col"))

# Enrichments
taxify_download_enrichment(c(
  "iucn",
  "zanne",
  "eive",
  "elton_traits"
))

# Now the analysis can run fully offline
result <- taxify(species_list, backend = c("wfo", "col"))
result <- add_iucn(result)
result <- add_zanne(result)
```

In a CI/CD or cluster environment, the download step can run in a setup
script or container build phase. The matching step then operates entirely
from local disk, with no network dependency and no risk of mid-run download
failures. This separation also makes the pipeline reproducible: the download
step pins a specific backbone version (recorded in the `meta.json` sidecar
file), and the matching step uses whatever version is on disk.

For a targets or drake plan, the download calls fit naturally as upstream
targets that the matching targets depend on. The return value (the `.vtr`
path) can be passed through the dependency graph, though in practice the
path is resolved internally by `ensure_backbone()` and does not need to be
passed explicitly.

To see which enrichments are available and their current versions:

```{r}
list_enrichments()
#>                name version    nrow static
#> 1 iucn 2026.04   59583  FALSE
#> 2               griis 2026.04   98131  FALSE
#> 3                wcvp 2026.04 1973234  FALSE
#> 4                eive     1.0   14835   TRUE
#> 5        elton_traits     1.0   15394   TRUE
#> 6              avonet     1.0   11009   TRUE
#> ...
```

Static enrichments (those based on published, version-locked datasets like
EltonTraits 1.0 or PanTHERIA 1.0) are never re-downloaded after the initial
fetch. Non-static enrichments (iucn, griis, wcvp,
common_names) are checked once per session and updated if a newer build is
available.


## Practical scaling guidance

The table below summarizes recommended settings by list size. These are
guidelines, not hard thresholds. The actual performance depends on input
cleanliness (how many names need fuzzy matching), backbone size (WFO vs.
GBIF), and hardware (number of cores, available RAM, disk speed).

**Under 1,000 names.** The defaults work well. `taxify(names)` with
`fuzzy = TRUE` and a single backend completes in a few seconds. No tuning
is needed. This is the regime for most interactive analysis: a field survey,
a thesis species list, a table extracted from a paper. Memory usage is
negligible.

**1,000 to 50,000 names.** If the input is clean (names from a curated
database, a previous taxify run, or a standard checklist), consider
`fuzzy = FALSE`. The exact pipeline handles case differences, Latin
orthographic variants (e.g., *-ii* vs. *-i* endings), and infraspecific-to-
species fallback without string-distance computation. Enabling fuzzy on a
clean list of 50,000 names might add 30-60 seconds for no practical gain.
If the input has known quality issues (OCR transcriptions, citizen science
data), leave fuzzy on and expect 1-3 minutes against WFO.

**50,000 to 500,000 names.** Backend ordering starts to matter. Put the
backbone that covers the dominant taxon group first. For a plant list, `"wfo"`
alone suffices. For mixed-kingdom lists, `c("wfo", "col")` resolves most
names on the faster WFO pass. Consider the two-pass pattern (exact first,
fuzzy on unmatched) if only a small fraction of names have quality issues.
The GBIF backbone at 6.4 million rows is the most expensive for fuzzy
matching; avoid it as the first backend unless the list is primarily
non-plant, non-marine taxa not covered by COL.

**Over 500,000 names.** Batch in chunks of 50,000-100,000 names. The
backbone stays cached across chunks, so there is no repeated initialization
cost. Write results to disk per chunk if total memory is a concern. Clear
the cache between analysis phases (matching, enrichment, downstream
modelling) to keep memory usage bounded. If enriching with multiple
layers, apply all enrichments to each chunk before writing rather than
accumulating the full result in memory and enriching after.

```{r}
# Pattern for 500,000+ names with enrichments
for (i in seq_along(chunks)) {
  res <- taxify(chunks[[i]], backend = "wfo",
                fuzzy = TRUE, verbose = FALSE)
  res <- add_iucn(res, verbose = FALSE)
  res <- add_zanne(res, verbose = FALSE)
  saveRDS(res, sprintf("results/chunk_%04d.rds", i))
}
```

**Backend selection cheat sheet:**

| List composition | Recommended backend(s) |
|-----------------|----------------------|
| Plants only | `"wfo"` |
| Plants + animals | `c("wfo", "col")` |
| Marine taxa | `c("worms", "col")` |
| Fungi | `c("fungorum", "col")` |
| Algae | `c("algaebase", "col")` |
| All kingdoms, maximum coverage | `c("wfo", "col", "gbif")` |
| Molecular/genomic taxa | `c("ncbi", "col")` |
| North American biodiversity | `c("itis", "col")` |

For any single-kingdom list, starting with the specialist backbone (WFO for
plants, WoRMS for marine, NCBI for molecular) and falling back to COL or
GBIF gives the best balance of speed and coverage. The specialist backbone
resolves most names quickly (smaller backbone, faster exact pass), and the
generalist backbone catches the remainder.


## The fuzzy_threshold parameter

The default fuzzy threshold is 0.2 (normalized Damerau-Levenshtein distance:
edits divided by the maximum of the two name lengths). A threshold of 0.2
allows roughly one edit per five characters, which catches single-character
typos in binomials of typical length (15-25 characters). This is a good
default for most use cases.

For large lists with noisy input (OCR, handwriting transcription), a
slightly higher threshold like 0.25 catches more misspellings but also
increases false positives. For clean input where fuzzy matching serves only
as a safety net, a lower threshold like 0.1 or 0.15 reduces the risk of
incorrect matches without sacrificing much recall.

The threshold also affects performance. A higher threshold means more
backbone entries pass the distance filter, which means more candidate
matches to evaluate and rank. The difference is modest for most inputs but
can be noticeable for very large genera: at threshold 0.2, a query against
*Astragalus* (~3,000 WFO entries) might return 5 candidates; at 0.3, it
might return 20.

An alternative mode uses integer thresholds. Setting `fuzzy_threshold = 2L`
allows at most 2 raw edit operations regardless of name length. This is
useful for long infraspecific names where a normalized threshold of 0.2
might allow too many edits. Integer thresholds are not supported with the
Jaro-Winkler method (`fuzzy_method = "jw"`).

```{r}
# Tight threshold for clean input
result <- taxify(clean_names, backend = "wfo",
                 fuzzy = TRUE, fuzzy_threshold = 0.1)

# Integer threshold: at most 2 edits, period
result <- taxify(noisy_names, backend = "wfo",
                 fuzzy = TRUE, fuzzy_threshold = 2L)
```


## Summary of performance-relevant functions

| Function | Purpose |
|----------|---------|
| `taxify(..., fuzzy = FALSE)` | Skip fuzzy matching for clean input |
| `taxify(..., backend = c("wfo", "col"))` | Multi-backend fallback chain |
| `taxify_data_dir()` | Find where backbones are stored |
| `taxify_download()` | Pre-download backbone `.vtr` files |
| `taxify_download_enrichment()` | Pre-download enrichment `.vtr` files |
| `taxify_clear_cache()` | Free backbone memory after matching |
| `taxify_refresh_manifest()` | Force re-check for backbone updates |
| `list_enrichments()` | See available enrichments and versions |
