---
title: "Migrating from taxize and WorldFlora"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from taxize and WorldFlora}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why migrate

The R ecosystem has two well-established packages for taxonomic name
resolution: [taxize](https://docs.ropensci.org/taxize/) for
multi-database API lookups, and
[WorldFlora](https://cran.r-project.org/package=WorldFlora) for offline
matching against the WFO backbone. Both work. If your workflow already
uses one of them and you are happy with it, there is no urgent reason to
switch.

That said, there are situations where taxify offers a better fit:

- **Multiple backbones.** taxify matches against seven backbones offline
  and can chain them in a single call:
  `taxify(names, backend = c("wfo", "col", "gbif"))`.
- **Speed at scale.** The matching engine is written in C with
  genus-blocked fuzzy joins. Ten thousand names resolve in seconds.
- **Enrichments.** Results pipe directly into twelve published trait and
  status datasets (IUCN, GRIIS, WCVP, EIVE, EltonTraits, etc.) with a
  single `|>` chain.
- **Reproducibility.** Backbones are versioned files on disk. The
  `backbone_version` column records exactly which snapshot was used.

This vignette maps the old APIs to their taxify equivalents, walks
through three side-by-side examples, and is honest about what taxify
does not cover.


## Function mapping: taxize to taxify

The table below maps the taxize name-resolution functions to their
closest taxify equivalent.

| taxize function | taxify equivalent | Notes |
|---|---|---|
| `gnr_resolve()` | `taxify()` | Any backend; returns best match per name |
| `classification()` | `taxify()` | `family`, `genus`, `rank` columns in the output; `add_col_info()` for full hierarchy |
| `synonyms()` | `taxify()` | `is_synonym` + `accepted_name` columns in the output |
| `tax_name()` | `taxify()` | `family`, `genus`, `rank` columns |
| `sci2comm()` | `add_common_names()` | Pipe enrichment; GBIF vernacular names by language |

taxize also has functions that serve a different purpose (fetching
database IDs, enumerating child taxa, retrieving occurrence or sequence
data). These are not name-resolution functions, so taxify does not cover
them. The "What taxify does not do" section below points to the right
packages for those tasks.

The key structural difference: taxize returned results in varied formats
depending on the function (`classification()` gave a nested list of
data.frames, `synonyms()` another nested list, `get_tsn()` a character
vector with attributes). taxify returns the same 16-column data.frame
from every call. Synonym status, classification, and match quality are
columns, not separate API calls.


## Function mapping: WorldFlora to taxify

| WorldFlora function | taxify equivalent | Notes |
|---|---|---|
| `WFO.match()` | `taxify(backend = "wfo")` | Exact + fuzzy in one call |
| `WFO.one()` | `taxify()` | Best-match selection is automatic |
| `WFO.match.fuzzyjoin()` | `taxify(fuzzy = TRUE)` | Enabled by default; genus-blocked Damerau-Levenshtein |
| `WFO.synonyms()` | `taxify()` | `is_synonym`, `accepted_name`, `accepted_id` in output |

WorldFlora returns a wide data.frame with WFO-specific column names
(`scientificName`, `taxonID`, `taxonomicStatus`, `acceptedNameUsageID`,
plus authorship and bibliographic fields). taxify normalizes these into
a backend-agnostic schema: `matched_name`, `taxon_id`, `accepted_name`,
`accepted_id`, and so on. The WFO-specific columns are still accessible
via `add_wfo_info()` when needed, but the default output is the same 16
columns whether the backend is WFO, COL, or GBIF.

taxify also handles backbone management automatically: the first
`taxify()` call downloads the backbone, subsequent calls reuse the local
copy, and a once-per-session version check keeps it current.


## Example 1: Basic name resolution

With taxize, name resolution typically meant several separate calls:
`gnr_resolve()` for matching, `get_gbifid()` for IDs,
`classification()` for hierarchy, `synonyms()` for synonym status.

```{r taxize-basic}
# --- taxize ---
library(taxize)

names <- c("Quercus robur", "Pinus sylvestris", "Betula pendula",
           "Panthera leo", "Salmo trutta")

resolved  <- gnr_resolve(names, best_match_only = TRUE)
gbif_ids  <- get_gbifid(names)
class_list <- classification(gbif_ids, db = "gbif")
syn_list   <- synonyms(gbif_ids, db = "gbif")
```

With taxify, all of that is one call:

```{r taxify-basic}
# --- taxify ---
library(taxify)

names <- c("Quercus robur", "Pinus sylvestris", "Betula pendula",
           "Panthera leo", "Salmo trutta")

result <- taxify(names, backend = "gbif")

result$accepted_name
result$family
result$genus
result$is_synonym
result$taxon_id        # GBIF usage key
```

The output is a data.frame with 16 columns and one row per input name.


## Example 2: WFO matching with fuzzy + synonyms

With WorldFlora, the typical workflow is: load the backbone, run exact
matching, apply fuzzy matching separately, then pick the best match.

```{r worldflora-fuzzy}
# --- WorldFlora ---
library(WorldFlora)

wfo_data <- read.delim("classification.txt")

names <- c("Quercus robur", "Quercus pedonculata",
           "Pinus silvestris", "Rosa canina")
exact <- WFO.match(names, WFO.data = wfo_data)
fuzzy <- WFO.match.fuzzyjoin(names, WFO.data = wfo_data)
best  <- WFO.one(fuzzy)
```

With taxify, exact matching, fuzzy matching, and synonym resolution
happen in a single call:

```{r taxify-fuzzy}
# --- taxify ---
library(taxify)

names <- c("Quercus robur", "Quercus pedonculata",
           "Pinus silvestris", "Rosa canina")

result <- taxify(names, backend = "wfo")

# Misspellings are caught by fuzzy matching:
result[, c("input_name", "matched_name", "match_type", "fuzzy_dist")]
#   input_name           matched_name        match_type fuzzy_dist
# 1 Quercus robur        Quercus robur       exact              NA
# 2 Quercus pedonculata  Quercus pedunculata fuzzy           0.053
# 3 Pinus silvestris     Pinus sylvestris    fuzzy           0.063
# 4 Rosa canina          Rosa canina         exact              NA

# Synonyms resolved automatically:
result[, c("input_name", "is_synonym", "accepted_name")]
```

`Quercus pedonculata` is both a misspelling and a synonym. taxify
handles both: the fuzzy matcher corrects the spelling to `Quercus
pedunculata`, and the synonym resolver maps it to `Quercus robur`.


## Example 3: Multi-backend fallback with enrichments

taxify can chain multiple backbones in a single call. Unmatched names
cascade to the next backbone automatically.

```{r taxify-fallback}
library(taxify)

# Mixed kingdom input: plants, animals, fungi
names <- c(
  "Quercus robur",         # plant (WFO primary)
  "Panthera leo",          # animal (not in WFO, picked up by GBIF)
  "Amanita muscaria",      # fungus (not in WFO, picked up by GBIF)
  "Salmo trutta",          # fish (not in WFO, picked up by GBIF)
  "Arabidopsis thaliana"   # plant (in both WFO and GBIF)
)

# WFO first (best for plants), GBIF as fallback (all kingdoms)
result <- taxify(names, backend = c("wfo", "gbif"))

# The backend column shows which database matched each name:
result[, c("input_name", "backend", "family")]
#   input_name            backend family
# 1 Quercus robur         wfo     Fagaceae
# 2 Panthera leo          gbif    Felidae
# 3 Amanita muscaria      gbif    Amanitaceae
# 4 Salmo trutta          gbif    Salmonidae
# 5 Arabidopsis thaliana  wfo     Brassicaceae

# Enrich with traits:
result |>
  add_conservation_status() |>
  add_woodiness()

# Or join custom data:
my_traits <- data.frame(
  species = c("Quercus robur", "Panthera leo"),
  max_height_m = c(35, NA),
  body_mass_kg = c(NA, 190)
)
result |> add_data(my_traits, species_col = "species")
```


## Key differences at a glance

**Offline matching.** taxify downloads backbone files once and matches
locally. After the initial download (typically 50–300 MB depending on
the backbone), no internet connection is needed.

**Multi-backend.** taxify supports seven backbones through a single
function, with optional fallback chains that cascade unmatched names
automatically.

**Output format.** taxify always returns a data.frame with 16
standardized columns, regardless of the backend:

| Column | Type | Content |
|---|---|---|
| `input_name` | character | Original name as submitted |
| `matched_name` | character | Closest match in the backbone |
| `accepted_name` | character | Accepted name after synonym resolution |
| `taxon_id` | character | Backend-specific ID of the matched name |
| `accepted_id` | character | ID of the accepted name |
| `rank` | character | Taxonomic rank (species, genus, family, etc.) |
| `family` | character | Family name |
| `genus` | character | Genus name |
| `epithet` | character | Specific epithet |
| `authorship` | character | Taxonomic authority |
| `is_synonym` | logical | Was the matched name a synonym? |
| `is_hybrid` | logical | Hybrid marker detected in the input? |
| `match_type` | character | `"exact"`, `"exact_ci"`, `"fuzzy"`, or `"none"` |
| `fuzzy_dist` | numeric | Normalized edit distance (NA if exact) |
| `backend` | character | Which backend matched this name |
| `backbone_version` | character | Backend name, version, and download date |

**Speed.** taxify uses vectra's C-level join engine with hash indexes
and genus-blocked fuzzy joins, processing thousands of names per second.

**Reproducibility.** taxify pins backbone versions locally and records
the version string in the `backbone_version` column of every result. The
same backbone file produces the same output indefinitely. Version
pinning is also available: `taxify_download_vtr("wfo", version =
"2024.06")` downloads a specific release.


## What taxify does not do

taxify is a name matcher. It resolves scientific names to accepted
names, returns classification metadata, and joins enrichment layers.
Several things that taxize or other packages handle are outside its
scope.

**Common-to-scientific name lookup.** taxize had `comm2sci()` to go
from "European robin" to *Erithacus rubecula*. taxify matches scientific
names, not vernacular input. For that direction, the GBIF API
(`rgbif::name_suggest()`) accepts common names and returns candidates.

**Downstream taxa.** taxize's `downstream()` returned all children of a
higher taxon (e.g., all species in a genus). taxify does not enumerate
children. For tree-based queries, the rotl package provides access to
the Open Tree of Life synthetic tree, and rgbif's `name_usage()` can
list children of a GBIF usage key.

**Phylogenetic trees.** For phylogenetic data, use rotl (Open Tree of
Life) or phylomatic.

**Occurrence data.** For occurrence data, rgbif and spocc are the
standard tools.

**Sequence data.** For sequence retrieval, the rentrez package handles
GenBank/NCBI queries directly.

**Real-time API lookups.** By design, taxify queries local files. If a
name was added to a backbone yesterday and taxify's local copy is from
last month, taxify will not find it until the backbone is updated.
For workflows where freshness matters more than reproducibility, a
direct API client (rgbif, worrms, ritis) may be the better fit.


## When not to switch

taxify is not a universal replacement for every taxize use case. A few
situations where the other tools may be more appropriate:

- **Common-to-scientific lookups** (`comm2sci()`). If the starting point
  is vernacular names, the GBIF API (`rgbif::name_suggest()`) accepts
  common names and returns scientific name candidates.

- **Downstream taxa enumeration.** If the goal is to list all species in
  a family or all subspecies of a species, taxify does not provide that
  query. Use `rgbif::name_usage(key, data = "children")` or
  `rotl::tol_subtree()`.

- **Interactive, per-name resolution with manual disambiguation.** taxize
  had interactive modes where the user could pick among multiple
  candidates. taxify picks the best match automatically (accepted name
  over synonym, species rank over higher ranks, lowest ID as tiebreaker).
  If manual control over ambiguous matches is needed, direct API calls
  may be preferable.

- **Data freshness over reproducibility.** taxify's strength is
  reproducible, versioned, offline matching. If the priority is to
  always use the very latest backbone update (published hours ago),
  querying the API directly via rgbif, worrms, or ritis avoids the
  delay between upstream publication and taxify's next backbone release.


## Discovering available enrichments

taxify bundles 12 enrichment datasets that cover conservation status,
invasive species, functional traits, morphological measurements, and
vernacular names. These are joined to the taxify result by piping
through `add_*()` functions.

```{r list-enrichments}
# See all available enrichments and their metadata
list_enrichments()
```

Each enrichment downloads automatically on first use and is cached
locally, following the same pattern as backbones. The full list:
`add_conservation_status()`, `add_invasive_status()`, `add_wcvp()`,
`add_eive()`, `add_elton_traits()`, `add_avonet()`, `add_pantheria()`,
`add_amphibio()`, `add_common_names()`, `add_woodiness()`,
`add_diaz_traits()`, and `add_leda()`.


## Summary

Migrating from taxize or WorldFlora to taxify means replacing multiple
function calls with `taxify()` plus optional `add_*()` pipes. The
output is a flat data.frame, not nested lists. Matching runs offline
against versioned backbone files, so results do not change between
sessions unless the user explicitly updates the backbone.

For things taxify does not handle (downstream taxa, occurrence data,
phylogenetic trees, sequence retrieval), the specialized packages (rotl,
rgbif, spocc, rentrez) remain the right tools. taxify handles the
name-matching step that comes before all of those.
