---
title: "Migrating from taxize, WorldFlora, and related tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from taxize, WorldFlora, and related tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## The taxonomic-resolution landscape in R

The R ecosystem has a rich set of taxonomic name-resolution tools. Each
takes a different design choice along three axes: where the data lives
(local files or remote APIs), how many backbones are bundled, and what
the package returns. The table below summarizes the options most likely
to overlap with a taxify workflow.

| Package | Source data | Coverage | Access | Closest taxify analogue |
|---|---|---|---|---|
| [taxize](https://docs.ropensci.org/taxize/) | ~20 web services (NCBI, ITIS, GBIF, EOL, IUCN, WoRMS, Tropicos, ...) | All kingdoms | Live API | `taxify(backend = c(...))` with the relevant local backbone(s) |
| [WorldFlora](https://cran.r-project.org/package=WorldFlora) | World Flora Online classification (`WFO.match`) | Land plants (vascular + bryophytes) | Local file | `taxify(backend = "wfo")` |
| [lcvplants](https://github.com/idiv-biodiversity/lcvplants) | Leipzig Catalogue of Vascular Plants | Vascular plants | Bundled in package | `taxify(backend = "lcvp")` |
| [rWCVP](https://matildabrown.github.io/rWCVP/) | World Checklist of Vascular Plants (Kew) | Vascular plants | Local snapshot | `taxify(backend = "wcvp")` |
| [taxadb](https://docs.ropensci.org/taxadb/) | GBIF, ITIS, COL, NCBI, OTT, WFO snapshots | All kingdoms | Local DuckDB / MonetDB | `taxify(backend = c(...))` |
| [Taxonstand](https://cran.r-project.org/package=Taxonstand) | The Plant List (retired by Kew in 2013, superseded by WCVP and WFO) | Vascular plants | Bundled in package | `taxify(backend = c("wcvp", "wfo"))` |
| [U.Taxonstand](https://github.com/ecoinfor/U.Taxonstand) | User-supplied or bundled checklists | Configurable | Local | `taxify(backend = ...)` plus `add_data()` |
| [bdc](https://brunobrr.github.io/bdc/) | taxadb + GNR for the taxonomic step inside a larger biodiversity-cleaning workflow | All kingdoms | Local + API | `taxify()` for the matching step |
| [TNRS](https://cran.r-project.org/package=TNRS) | TNRS web service (BIEN / iDigBio) | Plants | Live API | `taxify(backend = "wfo")` or similar |
| [rgbif](https://docs.ropensci.org/rgbif/), [worrms](https://docs.ropensci.org/worrms/), [ritis](https://docs.ropensci.org/ritis/) | GBIF / WoRMS / ITIS web APIs | One backbone each | Live API | `taxify(backend = "gbif" / "worms" / "itis")` |

If your workflow already uses one of these and you are happy with it,
there is no urgent reason to switch.

That said, there are situations where taxify offers a better fit:

- **Multiple backbones.** taxify matches against seven backbones offline
  and can chain them in a single call:
  `taxify(names, backend = c("wfo", "col", "gbif"))`.
- **Speed at scale.** The matching engine is written in C with
  genus-blocked fuzzy joins. Ten thousand names resolve in seconds.
- **Enrichments.** Results pipe directly into more than sixty published trait
  and status datasets (IUCN, GRIIS, WCVP, EIVE, EltonTraits, etc.) with a
  single `|>` chain.
- **Reproducibility.** Backbones are versioned files on disk. The
  `backbone_version` column records exactly which snapshot was used.

This vignette maps the old APIs to their taxify equivalents, walks
through three side-by-side examples, and is honest about what taxify
does not cover.


## Function mapping: taxize to taxify

The table below maps the taxize name-resolution functions to their
closest taxify equivalent.

| taxize function | taxify equivalent | Notes |
|---|---|---|
| `gnr_resolve()` | `taxify()` | Any backend; returns best match per name |
| `classification()` | `taxify()` | `family`, `genus`, `rank` columns in the output; `add_col_info()` for full hierarchy |
| `synonyms()` | `taxify()` | `is_synonym` + `accepted_name` columns in the output |
| `tax_name()` | `taxify()` | `family`, `genus`, `rank` columns |
| `sci2comm()` | `add_common_names()` | Pipe enrichment; GBIF vernacular names by language |

taxize also has functions that serve a different purpose (fetching
database IDs, enumerating child taxa, retrieving occurrence or sequence
data). These are not name-resolution functions, so taxify does not cover
them. The "What taxify does not do" section below points to the right
packages for those tasks.

The key structural difference: taxize returned results in varied formats
depending on the function (`classification()` gave a nested list of
data.frames, `synonyms()` another nested list, `get_tsn()` a character
vector with attributes). taxify returns the same 16-column data.frame
from every call. Synonym status, classification, and match quality are
columns, not separate API calls.


## Function mapping: WorldFlora to taxify

| WorldFlora function | taxify equivalent | Notes |
|---|---|---|
| `WFO.match()` | `taxify(backend = "wfo")` | Both do exact + fuzzy in one call. `WFO.match()` returns several candidate rows per input name |
| `WFO.one()` | `taxify()` | The selection step. `WFO.one()` collapses `WFO.match()`'s candidate rows to one best match per input; `taxify()` returns one best-match row directly, so the two steps are one call |
| `WFO.match.fuzzyjoin()` | `taxify(backend = "wfo")` | Same matching as `WFO.match()`, faster `fuzzyjoin`-based engine; taxify is fast by default, so there is no separate call |
| `WFO.synonyms(accepted)` | -- | Reverse direction: `WFO.synonyms()` expands an accepted name to its synonyms (one-to-many). taxify resolves the other way, synonym to accepted (`is_synonym`, `accepted_name`, `accepted_id`), so there is no direct equivalent |

WorldFlora returns a wide data.frame with WFO-specific column names
(`scientificName`, `taxonID`, `taxonomicStatus`, `acceptedNameUsageID`,
plus authorship and bibliographic fields). taxify normalizes these into
a backend-agnostic schema: `matched_name`, `taxon_id`, `accepted_name`,
`accepted_id`, and so on. The WFO-specific columns are still accessible
via `add_wfo_info()` when needed, but the default output is the same 16
columns whether the backend is WFO, COL, or GBIF.

taxify also handles backbone management automatically: the first
`taxify()` call downloads the backbone, subsequent calls reuse the local
copy, and a once-per-session version check keeps it current.


## Function mapping: lcvplants to taxify

[lcvplants](https://github.com/idiv-biodiversity/lcvplants) wraps the
Leipzig Catalogue of Vascular Plants and ships the LCVP table as bundled
data. The package centres on `LCVP()` and `lcvp_search()`.

| lcvplants function | taxify equivalent | Notes |
|---|---|---|
| `LCVP(splist)` | `taxify(splist, backend = "lcvp")` | Returns the standardized 16-column data.frame |
| `lcvp_search()` | `taxify()` | Search by name; same output schema |
| `lcvp_fuzzy_search()` | `taxify(fuzzy = TRUE)` | Genus-blocked Damerau-Levenshtein; on by default |
| `tab_lcvp` (data object) | `taxify_data_dir() / lcvp / latest / lcvp.vtr` | The LCVP snapshot is shipped as a `.vtr` file rather than an in-package data object |

The LCVP and WCVP backbones can be combined in a single fallback chain
to arbitrate between the Leipzig and Kew vascular-plant authorities:

```{r taxify-lcvp-wcvp}
result <- taxify(plant_names, backend = c("wcvp", "lcvp", "wfo"))
result[, c("input_name", "accepted_name", "backend")]
```


## Function mapping: rWCVP to taxify

[rWCVP](https://matildabrown.github.io/rWCVP/) is the Kew package for
the World Checklist of Vascular Plants. Its name-resolution side centres
on `wcvp_match_names()` and `wcvp_check_gbif()`; it also has a strong
distribution-query side that taxify does not replace.

| rWCVP function | taxify equivalent | Notes |
|---|---|---|
| `wcvp_match_names()` | `taxify(backend = "wcvp")` | Exact + fuzzy in one call |
| `wcvp_check_gbif()` | `taxify(backend = c("wcvp", "gbif"))` | Cascade WCVP first, GBIF as fallback |
| `wcvp_distribution()` | `add_wcvp()` | Native range by TDWG region (the `add_wcvp()` enrichment) |
| `wcvp_synonyms()` | `taxify()` | `is_synonym` and `accepted_name` columns in the output |
| `get_wcvp()` | automatic | The backbone downloads on first `taxify(backend = "wcvp")` call |

rWCVP's distribution-query functions (`wcvp_occ_mat()`, `generate_checklist()`)
operate on TDWG geography and are outside taxify's scope. For native-range
data joined to a name-resolved result, `add_wcvp()` covers the most common
case; for full geographic queries, rWCVP remains the right tool.


## Function mapping: taxadb to taxify

[taxadb](https://docs.ropensci.org/taxadb/) is the closest functional
analogue to taxify. Both store backbone snapshots locally and avoid
network calls at query time. The two packages differ in matching
strategy and integration: taxadb returns a long-format table for
exact-key joins, while taxify returns a flat one-row-per-input result
with fuzzy matching, synonym resolution, and trait enrichment built in.

| taxadb function | taxify equivalent | Notes |
|---|---|---|
| `td_create("itis")` | automatic | First `taxify(backend = "itis")` call downloads the `.vtr` snapshot |
| `filter_name(names, "itis")` | `taxify(names, backend = "itis")` | Exact match against the local snapshot |
| `filter_id(ids, "itis")` | not exposed | Use `vectra::tbl()` directly on the `.vtr` if needed |
| `synonyms(names, "itis")` | `taxify()` | `is_synonym`, `accepted_name`, `accepted_id` in the output |
| `clean_names()` | automatic | `taxify()` runs the cleaning pipeline (authorship, qualifiers, hybrid markers, orthography) before matching |
| (no fuzzy match) | `taxify(fuzzy = TRUE)` | Genus-blocked Damerau-Levenshtein, on by default |

The two largest practical differences:

- **Matching scope.** taxadb is built around exact lookups against
  pre-cleaned input. taxify cleans the input automatically and runs
  fuzzy matching on names that do not match exactly, which catches
  typos, orthographic variants, and authorship strings without a
  separate preprocessing step.
- **Output shape.** taxadb returns multiple rows per input when a name
  has multiple matches (you pick the row you want with `dplyr::filter`).
  taxify returns one row per input with a best-match selection rule
  (ACCEPTED over SYNONYM, species rank over higher ranks, lowest ID as
  tiebreaker), and reports the match type and fuzzy distance as columns.

For workflows that already use taxadb's column-oriented querying for
custom analyses, taxadb's approach is a clean fit. For workflows that
need a single resolved name per input plus enrichment joins, taxify's
flat output is closer to the goal.


## Function mapping: Taxonstand to taxify

[Taxonstand](https://cran.r-project.org/package=Taxonstand) was built
around The Plant List, which Kew retired in 2013 in favour of WCVP and
WFO. The package still works, but the underlying taxonomy has not been
updated since the retirement.

| Taxonstand function | taxify equivalent | Notes |
|---|---|---|
| `TPL(splist)` | `taxify(splist, backend = c("wcvp", "wfo"))` | Replace TPL with its successors |
| `TPLck()` | `taxify()` | Single-name check; same output schema |

The simplest migration is to replace `backend = "tpl"` with
`backend = c("wcvp", "wfo")` (or `backend = c("lcvp", "wcvp", "wfo")`
for triple-arbitration across the three large vascular-plant
authorities).


## Example 1: Basic name resolution

With taxize, name resolution typically meant several separate calls:
`gnr_resolve()` for matching, `get_gbifid()` for IDs,
`classification()` for hierarchy, `synonyms()` for synonym status.

```{r taxize-basic}
# --- taxize ---
library(taxize)

names <- c("Quercus robur", "Pinus sylvestris", "Betula pendula",
           "Panthera leo", "Salmo trutta")

resolved  <- gnr_resolve(names, best_match_only = TRUE)
gbif_ids  <- get_gbifid(names)
class_list <- classification(gbif_ids, db = "gbif")
syn_list   <- synonyms(gbif_ids, db = "gbif")
```

With taxify, all of that is one call:

```{r taxify-basic}
# --- taxify ---
library(taxify)

names <- c("Quercus robur", "Pinus sylvestris", "Betula pendula",
           "Panthera leo", "Salmo trutta")

result <- taxify(names, backend = "gbif")

result$accepted_name
result$family
result$genus
result$is_synonym
result$taxon_id        # GBIF usage key
```

The output is a data.frame with 16 columns and one row per input name.


## Example 2: WFO matching with fuzzy + synonyms

With WorldFlora, the typical workflow is: load the backbone, run the
match (exact and fuzzy together, here with the faster `fuzzyjoin` engine),
then collapse the candidate rows to one best match per input.

```{r worldflora-fuzzy}
# --- WorldFlora ---
library(WorldFlora)

wfo_data <- read.delim("classification.txt")

names <- c("Quercus robur", "Quercus pedonculata",
           "Pinus silvestris", "Rosa canina")
matched <- WFO.match.fuzzyjoin(names, WFO.data = wfo_data)
best    <- WFO.one(matched)
```

With taxify, exact matching, fuzzy matching, and synonym resolution
happen in a single call:

```{r taxify-fuzzy}
# --- taxify ---
library(taxify)

names <- c("Quercus robur", "Quercus pedonculata",
           "Pinus silvestris", "Rosa canina")

result <- taxify(names, backend = "wfo")

# Misspellings are caught by fuzzy matching:
result[, c("input_name", "matched_name", "match_type", "fuzzy_dist")]
#   input_name           matched_name        match_type fuzzy_dist
# 1 Quercus robur        Quercus robur       exact              NA
# 2 Quercus pedonculata  Quercus pedunculata fuzzy           0.053
# 3 Pinus silvestris     Pinus sylvestris    fuzzy           0.063
# 4 Rosa canina          Rosa canina         exact              NA

# Synonyms resolved automatically:
result[, c("input_name", "is_synonym", "accepted_name")]
```

`Quercus pedonculata` is both a misspelling and a synonym. taxify
handles both: the fuzzy matcher corrects the spelling to `Quercus
pedunculata`, and the synonym resolver maps it to `Quercus robur`.


## Example 3: Multi-backend fallback with enrichments

taxify can chain multiple backbones in a single call. Unmatched names
cascade to the next backbone automatically.

```{r taxify-fallback}
library(taxify)

# Mixed kingdom input: plants, animals, fungi
names <- c(
  "Quercus robur",         # plant (WFO primary)
  "Panthera leo",          # animal (not in WFO, picked up by GBIF)
  "Amanita muscaria",      # fungus (not in WFO, picked up by GBIF)
  "Salmo trutta",          # fish (not in WFO, picked up by GBIF)
  "Arabidopsis thaliana"   # plant (in both WFO and GBIF)
)

# WFO first (best for plants), GBIF as fallback (all kingdoms)
result <- taxify(names, backend = c("wfo", "gbif"))

# The backend column shows which database matched each name:
result[, c("input_name", "backend", "family")]
#   input_name            backend family
# 1 Quercus robur         wfo     Fagaceae
# 2 Panthera leo          gbif    Felidae
# 3 Amanita muscaria      gbif    Amanitaceae
# 4 Salmo trutta          gbif    Salmonidae
# 5 Arabidopsis thaliana  wfo     Brassicaceae

# Enrich with traits:
result |>
  add_conservation_status() |>
  add_woodiness()

# Or join custom data:
my_traits <- data.frame(
  species = c("Quercus robur", "Panthera leo"),
  max_height_m = c(35, NA),
  body_mass_kg = c(NA, 190)
)
result |> add_data(my_traits, species_col = "species")
```


## Key differences at a glance

**Offline matching.** taxify downloads backbone files once and matches
locally. After the initial download (typically 50–300 MB depending on
the backbone), no internet connection is needed.

**Multi-backend.** taxify supports seven backbones through a single
function, with optional fallback chains that cascade unmatched names
automatically.

**Output format.** taxify always returns a data.frame with 16
standardized columns, regardless of the backend:

| Column | Type | Content |
|---|---|---|
| `input_name` | character | Original name as submitted |
| `matched_name` | character | Closest match in the backbone |
| `accepted_name` | character | Accepted name after synonym resolution |
| `taxon_id` | character | Backend-specific ID of the matched name |
| `accepted_id` | character | ID of the accepted name |
| `rank` | character | Taxonomic rank (species, genus, family, etc.) |
| `family` | character | Family name |
| `genus` | character | Genus name |
| `epithet` | character | Specific epithet |
| `authorship` | character | Taxonomic authority |
| `is_synonym` | logical | Was the matched name a synonym? |
| `is_hybrid` | logical | Hybrid marker detected in the input? |
| `match_type` | character | `"exact"`, `"exact_ci"`, `"fuzzy"`, or `"none"` |
| `fuzzy_dist` | numeric | Normalized edit distance (NA if exact) |
| `backend` | character | Which backend matched this name |
| `backbone_version` | character | Backend name, version, and download date |

**Speed.** taxify uses vectra's C-level join engine with hash indexes
and genus-blocked fuzzy joins, processing thousands of names per second.

**Reproducibility.** taxify pins backbone versions locally and records
the version string in the `backbone_version` column of every result. The
same backbone file produces the same output indefinitely. Version
pinning is also available: `taxify_download_vtr("wfo", version =
"2024.06")` downloads a specific release.


## What taxify does not do

taxify is a name matcher. It resolves scientific names to accepted
names, returns classification metadata, and joins enrichment layers.
Several things that taxize or other packages handle are outside its
scope.

**Common-to-scientific name lookup.** taxize had `comm2sci()` to go
from "European robin" to *Erithacus rubecula*. taxify matches scientific
names, not vernacular input. For that direction, the GBIF API
(`rgbif::name_suggest()`) accepts common names and returns candidates.

**Downstream taxa.** taxize's `downstream()` returned all children of a
higher taxon (e.g., all species in a genus). taxify does not enumerate
children. For tree-based queries, the rotl package provides access to
the Open Tree of Life synthetic tree, and rgbif's `name_usage()` can
list children of a GBIF usage key.

**Phylogenetic trees.** For phylogenetic data, use rotl (Open Tree of
Life) or phylomatic.

**Occurrence data.** For occurrence data, rgbif and spocc are the
standard tools.

**Sequence data.** For sequence retrieval, the rentrez package handles
GenBank/NCBI queries directly.

**Real-time API lookups.** By design, taxify queries local files. If a
name was added to a backbone yesterday and taxify's local copy is from
last month, taxify will not find it until the backbone is updated.
For workflows where freshness matters more than reproducibility, a
direct API client (rgbif, worrms, ritis) may be the better fit.


## When the other packages are the better choice

taxify is one tool among several. A few situations where the related
packages remain the right answer:

- **Distribution and range queries.** rWCVP exposes WCVP's TDWG-region
  geography directly through `wcvp_distribution()`,
  `wcvp_occ_mat()`, and `generate_checklist()`. taxify covers
  name-resolution and the most common native-range join through
  `add_wcvp()`, but full geographic queries belong in rWCVP.

- **Live API access to upstream databases.** taxize, rgbif, worrms, ritis,
  and TNRS query their backends in real time. If you need a name added to
  a backbone yesterday, or you want the latest annotation for a single
  taxon, these packages return that immediately. taxify works against
  the snapshot on disk and only sees changes when the backbone is
  updated.

- **Common-to-scientific lookups.** taxize had `comm2sci()` to go from
  "European robin" to *Erithacus rubecula*. taxify matches scientific
  names, not vernacular input. For that direction,
  `rgbif::name_suggest()` accepts common names and returns candidates.

- **Downstream taxa enumeration.** If the goal is to list all species in
  a family or all subspecies of a species, taxify does not provide that
  query. Use `rgbif::name_usage(key, data = "children")` or
  `rotl::tol_subtree()`.

- **Wider biodiversity-data cleaning.** [bdc](https://brunobrr.github.io/bdc/)
  wraps the entire data-cleaning workflow (coordinate cleaning, dataset
  merging, taxonomic harmonization, occurrence flagging). taxify can
  replace its taxonomic step alone if you prefer offline backbones over
  taxadb + GNR, but the rest of bdc's pipeline is outside taxify's scope.

- **Interactive, per-name resolution with manual disambiguation.** taxize
  had interactive modes where the user could pick among multiple
  candidates. taxify picks the best match automatically (accepted name
  over synonym, species rank over higher ranks, lowest ID as tiebreaker).
  If manual control over ambiguous matches is needed, direct API calls
  may be preferable.

- **Column-oriented querying of a backbone.** taxadb stores backbones in
  DuckDB / MonetDB and exposes them through dplyr verbs, which is a
  natural fit if your analysis is itself a SQL-style transformation of
  the backbone. taxify exposes the underlying `.vtr` files through
  vectra for this kind of work, but taxadb's dplyr surface is more
  ergonomic for custom queries.


## Discovering available enrichments

taxify bundles 12 enrichment datasets that cover conservation status,
invasive species, functional traits, morphological measurements, and
vernacular names. These are joined to the taxify result by piping
through `add_*()` functions.

```{r list-enrichments}
# See all available enrichments and their metadata
list_enrichments()
```

Each enrichment downloads automatically on first use and is cached
locally, following the same pattern as backbones. The full list:
`add_conservation_status()`, `add_invasive_status()`, `add_wcvp()`,
`add_eive()`, `add_elton_traits()`, `add_avonet()`, `add_pantheria()`,
`add_amphibio()`, `add_common_names()`, `add_woodiness()`,
`add_diaz_traits()`, and `add_leda()`.


## Summary

Migrating from taxize, WorldFlora, lcvplants, rWCVP, taxadb, or
Taxonstand to taxify means replacing the package's resolution call with
`taxify(backend = ...)` and optional `add_*()` enrichment pipes. The
output is a flat 16-column data.frame, not nested lists or long-format
join tables, and matching runs offline against versioned backbone files
so results do not change between sessions unless the user explicitly
updates the backbone.

For things taxify does not handle (distribution queries, downstream
taxa, occurrence data, phylogenetic trees, sequence retrieval, live
API freshness), the specialized packages (rWCVP, rgbif, rotl, spocc,
rentrez, worrms, ritis) remain the right tools. taxify covers the
name-matching step that comes before most of those.
