---
title: "Migrating from taxize, WorldFlora, and related tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from taxize, WorldFlora, and related tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette maps the name-resolution calls of taxize, WorldFlora, lcvplants,
rWCVP, taxadb and Taxonstand to their taxify equivalents, then runs the same
tasks side by side in the old package and in taxify. The R tools for taxonomic
name resolution differ in where the data lives (local files or remote APIs), how
many backbones they cover, and what they return. taxify matches against
<!-- manifest:backbone-count -->19<!-- /manifest:backbone-count --> backbones
offline and can chain them in a single call:
`taxify(names, backbone = c("col", "gbif", "itis"))`. The matching engine is
written in C with genus-blocked fuzzy joins, and ten thousand names resolve in
seconds. Results pipe directly into more than eighty published trait and status
datasets (IUCN, GRIIS, WCVP, EIVE, EltonTraits, etc.) with a single `|>` chain.
Backbones are versioned files on disk, and the `backbone_version` column records
exactly which snapshot was used. There is no urgent reason to switch a workflow that already uses one of the
other packages and works well.

1. **Map** each call of the old package to its taxify equivalent with the
   tables below.
2. **Resolve** the names with `taxify()`, naming the `backbone` the old package
   read.
3. **Chain** backbones by passing a vector to `backbone`; unmatched names
   cascade to the next one.
4. **Enrich** the result with the `add_*()` layers that `list_enrichments()`
   lists, or join a custom table with `add_data()`.

## Example

```{r, eval = TRUE, message = FALSE, warning = FALSE}
library(taxify)
```

### The packages compared

The table summarizes the packages most likely to overlap with a taxify
workflow.

| Package | Source data | Coverage | Access | Closest taxify analogue |
|---|---|---|---|---|
| [taxize](https://docs.ropensci.org/taxize/) | ~20 web services (NCBI, ITIS, GBIF, EOL, IUCN, WoRMS, Tropicos, ...) | All kingdoms | Live API | `taxify(backbone = c(...))` with the relevant local backbone(s) |
| [WorldFlora](https://cran.r-project.org/package=WorldFlora) | World Flora Online classification (`WFO.match`) | Land plants (vascular + bryophytes) | Local file | `taxify(backbone = "wfo")` |
| [lcvplants](https://github.com/idiv-biodiversity/lcvplants) | Leipzig Catalogue of Vascular Plants | Vascular plants | Bundled in package | `taxify(backbone = "lcvp")` |
| [rWCVP](https://matildabrown.github.io/rWCVP/) | World Checklist of Vascular Plants (Kew) | Vascular plants | Local snapshot | `taxify(backbone = "wcvp")` for names; `add_wcvp()` for native ranges |
| [taxadb](https://docs.ropensci.org/taxadb/) | GBIF, ITIS, COL, NCBI, OTT, WFO snapshots | All kingdoms | Local DuckDB / MonetDB | `taxify(backbone = c(...))` |
| [Taxonstand](https://cran.r-project.org/package=Taxonstand) | The Plant List (retired by Kew in 2013, superseded by WCVP and WFO) | Vascular plants | Bundled in package | `taxify(backbone = "wfo")` |
| [U.Taxonstand](https://github.com/ecoinfor/U.Taxonstand) | User-supplied or bundled checklists | Configurable | Local | `taxify(backbone = ...)` plus `add_data()` |
| [bdc](https://brunobrr.github.io/bdc/) | taxadb + GNR for the taxonomic step inside a larger biodiversity-cleaning workflow | All kingdoms | Local + API | `taxify()` for the matching step |
| [TNRS](https://cran.r-project.org/package=TNRS) | TNRS web service (BIEN / iDigBio) | Plants | Live API | `taxify(backbone = "wfo")` or similar |
| [rgbif](https://docs.ropensci.org/rgbif/), [worrms](https://docs.ropensci.org/worrms/), [ritis](https://docs.ropensci.org/ritis/) | GBIF / WoRMS / ITIS web APIs | One backbone each | Live API | `taxify(backbone = "gbif" / "worms" / "itis")` |

### From taxize

The taxize name-resolution functions and their closest taxify equivalents:

| taxize function | taxify equivalent | Notes |
|---|---|---|
| `gnr_resolve()` | `taxify()` | Any backbone; returns best match per name |
| `classification()` | `taxify()` + `add_classification()` | `family`, `genus`, `rank` in the core output; `add_classification()` fills kingdom / phylum / class / order |
| `synonyms()` | `synonyms()` | Lists every synonym that resolves to a name's accepted taxon |
| `children()` | `children()` | Lists the accepted taxa within a genus or family |
| `downstream()` | `downstream()` | Every accepted taxon at a target rank beneath a higher taxon (species by default) |
| `upstream()` | `upstream()` | The ancestors of a taxon, one row per rank |
| `tax_name()` | `taxify()` | `family`, `genus`, `rank` columns |
| `sci2comm()` | `sci2comm()` or `add_common_names()` | GBIF, NCBI and Open Tree vernacular names; `add_common_names()` is the pipe enrichment, by language |
| `comm2sci()` | `comm2sci()` | Common name to accepted scientific name(s), read from the same vernacular tables |
| `id2name()` | `id2name()` | Backbone ID (GBIF key, TSN, AphiaID) to name, rank and accepted name |
| `class2tree()` | `class2tree()` | Taxonomy tree from the classification, as Newick |
| `lowest_common()` | `lowest_common()` | Deepest rank a set of names shares |

Database IDs come back in the `taxon_id` and `accepted_id` columns of every
`taxify()` result, so the `get_*id()` family has no separate call. taxize also
retrieves occurrence and sequence data; those tasks are outside taxify's scope,
and [When another package fits better](#when-another-package-fits-better) points
to the packages for them.

taxize returned results in varied formats depending on the function
(`classification()` gave a nested list of data.frames, `synonyms()` another
nested list, `get_tsn()` a character vector with attributes). taxify returns the
same 27-column data.frame from every `taxify()` call, with synonym status,
classification, and match quality as columns.

### From WorldFlora

| WorldFlora function | taxify equivalent | Notes |
|---|---|---|
| `WFO.match()` | `taxify(backbone = "wfo")` | Both do exact + fuzzy in one call. `WFO.match()` returns several candidate rows per input name |
| `WFO.one()` | `taxify()` | The selection step. `WFO.one()` collapses `WFO.match()`'s candidate rows to one best match per input; `taxify()` returns one best-match row directly, so the two steps are one call |
| `WFO.match.fuzzyjoin()` | `taxify(backbone = "wfo")` | Same matching as `WFO.match()`, faster `fuzzyjoin`-based engine; taxify is fast by default, so there is no separate call |
| `WFO.synonyms(accepted)` | `synonyms(x, backbone = "wfo")` | Expands a name to the synonyms of its accepted taxon (one-to-many). The forward direction, synonym to accepted, is in the `taxify()` output (`is_synonym`, `accepted_name`, `accepted_id`) |

WorldFlora returns a wide data.frame with WFO-specific column names
(`scientificName`, `taxonID`, `taxonomicStatus`, `acceptedNameUsageID`, plus
authorship and bibliographic fields). taxify normalizes these into a
backbone-agnostic schema: `matched_name`, `taxon_id`, `accepted_name`,
`accepted_id`, and so on. The WFO-specific columns are still accessible via
`add_wfo_info()`, and the default output has the same 27 columns whether the
backbone is WFO, COL, or GBIF.

taxify also manages the backbone: the first `taxify()` call downloads it,
subsequent calls reuse the local copy, and a once-per-session version check
keeps it current.

### From lcvplants

[lcvplants](https://github.com/idiv-biodiversity/lcvplants) wraps the Leipzig
Catalogue of Vascular Plants and ships the LCVP table as bundled data. The
package centres on `LCVP()` and `lcvp_search()`.

| lcvplants function | taxify equivalent | Notes |
|---|---|---|
| `LCVP(splist)` | `taxify(splist, backbone = "lcvp")` | Returns the standardized 27-column data.frame |
| `lcvp_search()` | `taxify()` | Search by name; same output schema |
| `lcvp_fuzzy_search()` | `taxify(fuzzy = TRUE)` | Genus-blocked Damerau-Levenshtein; on by default |
| `tab_lcvp` (data object) | `taxify_data_dir() / lcvp / latest / lcvp.vtr` | The LCVP snapshot is shipped as a `.vtr` file instead of an in-package data object |

The LCVP and WCVP backbones can be combined in a single fallback chain to
arbitrate between the Leipzig and Kew vascular-plant authorities:

```{r taxify-lcvp-wcvp}
plant_names <- c("Quercus robur", "Pinus sylvestris", "Betula pendula")
result <- taxify(plant_names, backbone = c("wcvp", "lcvp", "wfo"))
result[, c("input_name", "accepted_name", "backbone")]
```

### From rWCVP

[rWCVP](https://matildabrown.github.io/rWCVP/) is the Kew package for the World
Checklist of Vascular Plants. Its name-resolution side centres on
`wcvp_match_names()` and `wcvp_check_gbif()`; its distribution-query side has no
taxify replacement.

| rWCVP function | taxify equivalent | Notes |
|---|---|---|
| `wcvp_match_names()` | `taxify(backbone = "wcvp")` | Exact + fuzzy in one call |
| `wcvp_check_gbif()` | `taxify(backbone = c("wcvp", "gbif"))` | Cascade WCVP first, GBIF as fallback |
| `wcvp_distribution()` | `add_wcvp()` | Native range by TDWG region (the `add_wcvp()` enrichment) |
| `wcvp_synonyms()` | `taxify()`, `synonyms(backbone = "wcvp")` | `is_synonym` and `accepted_name` columns in the `taxify()` output; `synonyms()` lists the synonyms of a taxon |
| `get_wcvp()` | automatic | The backbone downloads on first `taxify(backbone = "wcvp")` call |

rWCVP's distribution-query functions (`wcvp_occ_mat()`, `generate_checklist()`)
operate on TDWG geography and are outside taxify's scope. For native-range data
joined to a name-resolved result, `add_wcvp()` covers the most common case; for
full geographic queries, rWCVP remains the right tool.

### From taxadb

[taxadb](https://docs.ropensci.org/taxadb/) is the closest functional analogue
to taxify. Both store backbone snapshots locally and avoid network calls at
query time. taxadb returns a long-format table for exact-key joins, while taxify
returns a flat one-row-per-input result with fuzzy matching, synonym
resolution, and trait enrichment built in.

| taxadb function | taxify equivalent | Notes |
|---|---|---|
| `td_create("itis")` | automatic | First `taxify(backbone = "itis")` call downloads the `.vtr` snapshot |
| `filter_name(names, "itis")` | `taxify(names, backbone = "itis")` | Exact match against the local snapshot |
| `filter_id(ids, "itis")` | `id2name(ids, backbone = "itis")` | Name, rank, classification and accepted name for each ID |
| `synonyms(names, "itis")` | `taxify()` | `is_synonym`, `accepted_name`, `accepted_id` in the output |
| `clean_names()` | automatic | `taxify()` runs the cleaning pipeline (authorship, qualifiers, hybrid markers, orthography) before matching |
| (no fuzzy match) | `taxify(fuzzy = TRUE)` | Genus-blocked Damerau-Levenshtein, on by default |

The two largest practical differences are matching scope and output shape.
taxadb is built around exact lookups against pre-cleaned input; taxify cleans
the input and runs fuzzy matching on names that do not match exactly, which
catches typos, orthographic variants, and authorship strings without a separate
preprocessing step. taxadb returns multiple rows per input when a name has
multiple matches, and the wanted row is picked with `dplyr::filter`. taxify
returns one row per input with a best-match selection rule (ACCEPTED over
SYNONYM, species rank over higher ranks, lowest ID as tiebreaker), and reports
the match type and fuzzy distance as columns.

### From Taxonstand

[Taxonstand](https://cran.r-project.org/package=Taxonstand) was built around The
Plant List, which Kew retired in 2013 in favour of WCVP and WFO. The package
still works, but the underlying taxonomy has not been updated since the
retirement.

| Taxonstand function | taxify equivalent | Notes |
|---|---|---|
| `TPL(splist)` | `taxify(splist, backbone = c("wcvp", "wfo"))` | Replace TPL with its successors |
| `TPLck()` | `taxify()` | Single-name check; same output schema |

The simplest migration replaces `backbone = "tpl"` with
`backbone = c("wcvp", "wfo")`, or with `backbone = c("lcvp", "wcvp", "wfo")` for
triple arbitration across the three large vascular-plant authorities.

### Basic name resolution

With taxize, name resolution typically meant several separate calls:
`gnr_resolve()` for matching, `get_gbifid()` for IDs, `classification()` for
hierarchy, `synonyms()` for synonym status.

```{r taxize-basic}
# --- taxize ---
library(taxize)

names <- c("Quercus robur", "Pinus sylvestris", "Betula pendula",
           "Panthera leo", "Salmo trutta")

resolved  <- gnr_resolve(names, best_match_only = TRUE)
gbif_ids  <- get_gbifid(names)
class_list <- classification(gbif_ids, db = "gbif")
syn_list   <- synonyms(gbif_ids, db = "gbif")
```

With taxify, all of that is one call, returning a data.frame with 27 columns and
one row per input name:

```{r taxify-basic}
# --- taxify ---
names <- c("Quercus robur", "Pinus sylvestris", "Betula pendula",
           "Panthera leo", "Salmo trutta")

result <- taxify(names, backbone = "gbif")

result$accepted_name
result$family
result$genus
result$is_synonym
result$taxon_id        # GBIF usage key
```

### WFO matching with fuzzy matching and synonyms

With WorldFlora, the typical workflow loads the backbone, runs the match (exact
and fuzzy together, here with the faster `fuzzyjoin` engine), then collapses the
candidate rows to one best match per input.

```{r worldflora-fuzzy}
# --- WorldFlora ---
library(WorldFlora)

wfo_data <- read.delim("classification.txt")

names <- c("Quercus robur", "Quercus pedonculata",
           "Pinus silvestris", "Rosa canina")
matched <- WFO.match.fuzzyjoin(names, WFO.data = wfo_data)
best    <- WFO.one(matched)
```

In taxify, exact matching, fuzzy matching, and synonym resolution happen in a
single call:

```{r taxify-fuzzy}
# --- taxify ---
names <- c("Quercus robur", "Quercus pedonculata",
           "Pinus silvestris", "Rosa canina")

result <- taxify(names, backbone = "wfo")

# Misspellings are caught by fuzzy matching:
result[, c("input_name", "matched_name", "match_type", "fuzzy_dist")]
#   input_name           matched_name        match_type fuzzy_dist
# 1 Quercus robur        Quercus robur       exact              NA
# 2 Quercus pedonculata  Quercus pedunculata fuzzy           0.053
# 3 Pinus silvestris     Pinus sylvestris    fuzzy           0.063
# 4 Rosa canina          Rosa canina         exact              NA

# Synonyms resolved automatically:
result[, c("input_name", "is_synonym", "accepted_name")]
```

`Quercus pedonculata` is both a misspelling and a synonym. The fuzzy matcher
corrects the spelling to `Quercus pedunculata`, and the synonym resolver maps it
to `Quercus robur`.

### Multi-backbone fallback with enrichments

A vector of backbones forms a fallback chain: names left unmatched by one
backbone cascade to the next.

```{r taxify-fallback}
# Mixed kingdom input: plants, animals, fungi
names <- c(
  "Quercus robur",         # plant (matched by WFO)
  "Panthera leo",          # animal (not in WFO, picked up by GBIF)
  "Amanita muscaria",      # fungus (not in WFO, picked up by GBIF)
  "Salmo trutta",          # fish (not in WFO, picked up by GBIF)
  "Arabidopsis thaliana"   # plant (in both WFO and GBIF)
)

# WFO first (best for plants), GBIF as fallback (all kingdoms)
result <- taxify(names, backbone = c("wfo", "gbif"))

# The backbone column shows which database matched each name:
result[, c("input_name", "backbone", "family")]
#   input_name            backbone family
# 1 Quercus robur         wfo     Fagaceae
# 2 Panthera leo          gbif    Felidae
# 3 Amanita muscaria      gbif    Amanitaceae
# 4 Salmo trutta          gbif    Salmonidae
# 5 Arabidopsis thaliana  wfo     Brassicaceae

# Enrich with traits:
result |>
  add_iucn() |>
  add_zanne()

# Or join custom data:
my_traits <- data.frame(
  species = c("Quercus robur", "Panthera leo"),
  max_height_m = c(35, NA),
  body_mass_kg = c(NA, 190)
)
result |> add_data(my_traits, species_col = "species")
```

### Discovering enrichments

The enrichment datasets cover conservation status, invasive species, functional
traits, morphological measurements, and vernacular names, and join to the result
through the `add_*()` functions. `list_enrichments()` lists every source-named
`add_*()` door with its metadata:

```{r list-enrichments}
list_enrichments()
```

Each enrichment downloads on first use and is cached locally, following the same
pattern as backbones. The cross-source `add_trait()` verb gathers a single trait
across every source that carries it.

## Key differences

**Offline matching.** taxify downloads backbone files once and matches locally.
After the initial download (typically 50–300 MB depending on the backbone), no
internet connection is needed.

**Multi-backbone.** taxify supports
<!-- manifest:backbone-count -->19<!-- /manifest:backbone-count --> backbones
through a single function, with optional fallback chains that cascade unmatched
names.

**Output format.** `taxify()` returns a data.frame with 27 standardized columns,
regardless of the backbone:

| Column | Type | Content |
|---|---|---|
| `input_name` | character | Original name as submitted |
| `matched_name` | character | Closest match in the backbone |
| `accepted_name` | character | Accepted name after synonym resolution |
| `taxon_id` | character | Backend-specific ID of the matched name |
| `accepted_id` | character | ID of the accepted name |
| `rank` | character | Taxonomic rank (species, genus, family, etc.) |
| `family` | character | Family name |
| `genus` | character | Genus name |
| `epithet` | character | Specific epithet |
| `authorship` | character | Taxonomic authority |
| `accepted_authorship` | character | Authorship of the accepted name |
| `is_synonym` | logical | Was the matched name a synonym? |
| `taxonomic_status` | character | The matched record's status as the backbone writes it (`"ACCEPTED"`, `"SYNONYM"`, WFO's `"UNCHECKED"`, ...) |
| `is_hybrid` | logical | Hybrid marker detected in the input? |
| `match_type` | character | `"exact"`, `"exact_ci"`, `"abbrev"`, `"fuzzy"`, `"hybrid_formula"`, `"rank_fallback"`, `"basionym"`, or `"none"` |
| `fuzzy_dist` | numeric | Normalized edit distance (NA if exact) |
| `n_ids` | integer | Distinct accepted taxa the backbone files the name under |
| `accepted_ids` | character | All of those accepted IDs, pipe-separated, the pick first |
| `backbone` | character | Which backbone matched this name |
| `backbone_version` | character | Backend name, version, and download date |
| `kingdom_group` | character | Coarse kingdom group (from the genus register) |
| `taxon_group` | character | Taxon grouping (from the genus register) |
| `life_form` | character | Life form (from the genus register) |
| `qualifier` | character | Canonical taxonomic qualifier (cf., aff., agg., s.l., ...) |
| `qualifier_position` | character | `"genus"` or `"species"` placement of the qualifier |
| `aggregate_fallback` | logical | TRUE when an aggregate query fell back to the binomial |
| `hybrid_type` | character | Hybrid formula / type detected |

**Speed.** taxify uses vectra's C-level join engine with hash indexes and
genus-blocked fuzzy joins, processing thousands of names per second.

**Reproducibility.** taxify pins backbone versions locally and records the
version string in the `backbone_version` column of every result. The same
backbone file produces the same output indefinitely. Version pinning is also
available: `taxify_download("col", version = "2024.06")` downloads a specific
release.

## When another package fits better

taxify resolves scientific names to accepted names, returns classification
metadata, and joins enrichment layers. The related packages remain the right
answer in these situations:

- **Distribution and range queries.** rWCVP exposes WCVP's TDWG-region
  geography directly through `wcvp_distribution()`, `wcvp_occ_mat()`, and
  `generate_checklist()`. taxify covers the most common native-range join
  through `add_wcvp()`; full geographic queries belong in rWCVP.

- **Live API access to upstream databases.** taxize, rgbif, worrms, ritis, and
  TNRS query their backbones in real time and return the latest annotation for a
  taxon immediately. taxify queries local files: if a name was added to a
  backbone yesterday and taxify's local copy is from last month, taxify will not
  find it until the backbone is updated. For workflows where freshness matters
  more than reproducibility, a direct API client (rgbif, worrms, ritis) may be
  the better fit. The same applies to common names: `comm2sci()` reads the
  bundled GBIF, NCBI and Open Tree vernacular tables, while
  `rgbif::name_suggest()` accepts common names against the live GBIF API.

- **Tree-based and phylogenetic queries.** `downstream()` and `children()` list
  descendants from the backbone's classification, and `class2tree()` builds a
  taxonomy tree from it. For the Open Tree of Life synthetic tree, use rotl
  (`rotl::tol_subtree()`); rgbif's `name_usage(key, data = "children")` lists
  the children of a GBIF usage key live. For phylogenetic data, use rotl or
  phylomatic.

- **Occurrence data.** rgbif and spocc are the standard tools.

- **Sequence data.** The rentrez package handles GenBank/NCBI queries directly.

- **Wider biodiversity-data cleaning.** [bdc](https://brunobrr.github.io/bdc/)
  wraps the entire data-cleaning workflow (coordinate cleaning, dataset merging,
  taxonomic harmonization, occurrence flagging). taxify can replace its
  taxonomic step alone when offline backbones are preferred over taxadb + GNR; the
  rest of bdc's pipeline is outside taxify's scope.

- **Interactive, per-name resolution with manual disambiguation.** taxize had
  interactive modes where the user could pick among multiple candidates. taxify
  picks the best match automatically (on GBIF a key with occurrence records
  over one without, then accepted name over synonym, species rank over higher ranks,
  lowest ID as tiebreaker), records every accepted ID of the name in
  `accepted_ids`, and lists them with `taxify_ids()`. If per-name interactive
  choice is needed, direct API calls may be preferable.

- **Column-oriented querying of a backbone.** taxadb stores backbones in
  DuckDB / MonetDB and exposes them through dplyr verbs, a natural fit when the
  analysis is itself a SQL-style transformation of the backbone. taxify exposes
  the underlying `.vtr` files through vectra for this kind of work, but taxadb's
  dplyr surface is more ergonomic for custom queries. For workflows that need a
  single resolved name per input plus enrichment joins, taxify's flat output is
  closer to the goal.
