---
title: "Constraining matches to a geographic region"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constraining matches to a geographic region}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## The problem

Fuzzy matching corrects a typo by finding the nearest real name. Most of the
time the nearest name is the one the recorder meant, but two species can sit a
single edit apart while living on different continents. A recorder working in
Belgium who writes a slightly misspelled name meant a Belgian plant, not its
one-letter neighbour from New Zealand. The string distance alone cannot tell
the two apart; the geography can.

`taxify()` takes a `region` argument for exactly this. When you set it,
`taxify()` prefers the fuzzy candidates that actually occur where you work and
sets the others aside. It never touches an exact match, so declaring a region
only ever changes which spelling correction wins, never a name that was already
right.

```{r}
library(taxify)

# a small regional list, with a couple of misspellings to correct
field_names <- c(
  "Gentiana acaulis", "Primula veris", "Pulsatilla vulgaris",
  "Gentiana acaulary", "Primula elatour"
)
```

## How the constraint works

The filter rests on two range sources. For vascular plants it uses WCVP, the
World Checklist of Vascular Plants, which records where each accepted species
occurs by TDWG botanical region. For marine taxa it uses WoRMS distribution
records rolled up to Marine Ecoregions of the World. `taxify()` resolves your
`region` input to codes in whichever vocabulary it belongs to, looks the
candidate fuzzy names up in the matching source, and drops an out-of-region
candidate when a better one survives. Three rules keep it conservative:

- It filters fuzzy candidates only. It trusts an exact or case-folded match as
  given.
- It never drops a candidate with no range data. Absence of data is not absence
  from the region, so a match neither source covers passes through untouched.
- It keeps every candidate for a name when all of them are out of region. The
  filter refines a match; it does not refuse one.

So the constraint is a soft preference. It breaks ties toward local species
and otherwise stays out of the way.

## By region name

The clearest input is a name. The bundled WGSRPD crosswalk accepts botanical
regions at three levels, so a country, a sub-continental region, or a continent
all work, case- and accent-insensitively.

```{r}
taxify("Gentiana acaulis", region = "Europe")
```

```
#>         input_name    accepted_name       family match_type fuzzy_dist backend
#> 1 Gentiana acaulis Gentiana acaulis Gentianaceae      exact         NA     WFO
```

The exact match comes back unchanged, since the region never touches one. The
constraint earns its keep on the fuzzy names in the same call: when a typo has
two corrections a single edit apart and only one of them grows in Europe, the
European one wins the tie. A name with no such conflict resolves exactly as it
would without a region.

`"Europe"` is a Level 1 region and expands to every European code; `"Middle
Europe"` is a Level 2 region; `"Belgium"` is a single Level 3 country. You can
pass several, and they union:

```{r}
taxify(field_names, region = c("Belgium", "Netherlands", "Germany"))
```

A three-letter token is read as a TDWG code directly, so `region = "BGM"`
(Belgium) and `region = "Belgium"` reach the same place. An unrecognised region
is dropped with a warning rather than failing the call, and a code that matches
no WCVP record simply makes the filter a no-op, so a typo in the region
degrades gracefully instead of producing wrong matches.

## By coordinates

When the data carry coordinates, hand them over directly. A point is mapped to
its botanical region by point-in-polygon against the WGSRPD Level 3 boundaries,
and the resulting codes are used the same way a region name would be.

```{r}
# Brussels: c(longitude, latitude)
taxify(field_names, coords = c(4.35, 50.85))
```

The order is `c(lon, lat)`. A single point, a two-column matrix or data.frame
of points, or a point-geometry spatial object all work; an `sf` object or a
terra `SpatVector` is reprojected to longitude/latitude on the way in. Points
and a `region` name can be combined, and their regions union.

```{r}
occ <- data.frame(
  lon = c(4.35, 5.12, 4.40),
  lat = c(50.85, 51.21, 50.50)
)
taxify(field_names, coords = occ)
```

The boundary file downloads once and stays cached. By default the lookup runs a
native ray-casting test, so no spatial package is required. With terra or sf
installed taxify uses that instead, which is faster on large point sets, and
`options(taxify.pip_engine = "terra" | "sf" | "native")` forces the choice.

## Native, introduced, or present

By default any WCVP record counts as in-region, native or introduced alike.
The `range` argument narrows that.

```{r}
# only count regions where WCVP lists the species as native
taxify(field_names, region = "Europe", range = "native")

# only introduced occurrences
taxify(field_names, region = "Europe", range = "introduced")
```

`range = "present"` is the default and the most permissive. `"native"` is
stricter and suits work that should ignore naturalised populations; a species
present in your region only as an introduction will not satisfy it, and its
out-of-region native correction can lose the tie. `"introduced"` is the mirror
image, for invasion work that wants the alien records specifically. The
argument is ignored when no region is set.

## Looking up regions

`taxify_regions()` returns the regions `region=` accepts so you can find the
right code or confirm a name resolves. With no argument it lists both
vocabularies; with a search term it filters, matching the code and all three
level names. The `scheme` column names the vocabulary a row belongs to, and
`scheme=` restricts the listing to one of them.

```{r}
taxify_regions("Belgium")
```

```
#>   code    name   level2_name level1_name scheme
#> 1  BGM Belgium Middle Europe      EUROPE wgsrpd
```

```{r}
# every code Europe expands to
nrow(taxify_regions("Europe"))
#> [1] 41

# browse the full table
head(taxify_regions())
```

The same crosswalk powers `add_wcvp()`, so the botanical codes here are the ones
that appear in native-range enrichment output.

## Marine regions

Marine names use the second range source: WoRMS distribution records rolled up
to Marine Ecoregions of the World. It arrives with the `marine_distribution`
asset, which downloads on first use, and from then on `region=` takes MEOW names
and codes the same way it takes botanical ones.

```{r}
taxify(c("Carcinus maenus", "Gadus morhua"), region = "North Sea")
```

```
#>        input_name   accepted_name     family match_type backend
#> 1 Carcinus maenus Carcinus maenas Carcinidae      fuzzy     col
#> 2    Gadus morhua    Gadus morhua    Gadidae      exact     col
```

An ecoregion name resolves to its own code. A province or realm name expands to
its member ecoregions, the way a TDWG Level 1 continent expands to its Level 3
codes.

```{r}
head(taxify_regions("Temperate Northern Atlantic", scheme = "meow"))
```

```
#>    code                           name                       level2_name
#> 1 25042                     Carolinian Warm Temperate Northwest Atlantic
#> 2 20043        Northern Gulf of Mexico Warm Temperate Northwest Atlantic
#> 3 25040     Gulf of Maine/Bay of Fundy Cold Temperate Northwest Atlantic
#> 4 20041                      Virginian Cold Temperate Northwest Atlantic
#> 5 20030                   Adriatic Sea                 Mediterranean Sea
#> 6 25033 Tunisian Plateau/Gulf of Sidra                 Mediterranean Sea
#>                   level1_name scheme
#> 1 Temperate Northern Atlantic   meow
#> 2 Temperate Northern Atlantic   meow
#> 3 Temperate Northern Atlantic   meow
#> 4 Temperate Northern Atlantic   meow
#> 5 Temperate Northern Atlantic   meow
#> 6 Temperate Northern Atlantic   meow
```

Coordinates behave the same way. A point at sea maps to the MEOW ecoregion
containing it, and `taxify()` unions that with the botanical lookup, so a single
`coords` argument serves a list holding both plants and marine animals.

```{r}
# central North Sea
taxify(field_names, coords = c(3.0, 56.0))
```

MEOW maps coastal and shelf waters, so a point over a deep ocean basin belongs
to no ecoregion. The filter treats that as absence of data and leaves those
names alone.

## What it covers, and what it does not

The two sources cover vascular plants and marine taxa. A name outside both
scopes has no range records, so the filter leaves it alone by design, which is
why a mixed list can carry a region without harming the matches it has no data
for.

Marine ranges inherit the grain of the underlying record. WoRMS files a
distribution against a named locality, and those run from a single bay to a
whole ocean basin; a record against a basin maps to every ecoregion inside it.
The median species spans 4 ecoregions and a quarter span exactly one, so most
ranges are sharp. About 0.3% span more than half the ocean, and for those the
filter is permissive: a species recorded that widely is in region wherever you
ask, so it breaks no ties. The constraint also acts on fuzzy candidates only, so it changes nothing
for a list that matches exactly throughout. It is most useful on regional field
lists with the usual crop of misspellings, where the right correction and a
plausible wrong one are a single edit apart.

The related check in `inspect()` looks at the other end of the pipeline. Rather
than steering a correction, it takes matched names and flags the ones WCVP does
not record in your region, surfacing a real but geographically out-of-place
species for review. The two share the `region`, `coords`, and `range`
arguments. See the
[name inspection vignette](https://gillescolling.com/taxify/articles/inspecting-names.html)
for that pass.

## Where to go next

- [Inspecting a name list](https://gillescolling.com/taxify/articles/inspecting-names.html)
  for the geographic outlier check that uses the same region inputs.

- [Fuzzy matching](https://gillescolling.com/taxify/articles/fuzzy-matching.html)
  for the candidate generation the region filter refines.

- [Enrichments](https://gillescolling.com/taxify/articles/enrichments.html)
  for `add_wcvp()`, which attaches native range on the same TDWG codes.
```
