---
title: "Databases: every backbone, enrichment, and trait"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)
library(taxify)

# Render from the manifest that ships with this build, so the tables always match
# the installed package and need no network. list_backbones() / list_enrichments()
# read the session manifest cache; point it at the bundled copy.
mp <- system.file("manifest.json", package = "taxify")
if (nzchar(mp)) {
  bundled <- jsonlite::read_json(mp, simplifyVector = FALSE)
  assign("manifest", bundled, envir = taxify:::.taxify_env)
}

link <- function(url) {
  # Some source_url values are bare DOIs (10.xxxx/...) rather than full URLs;
  # render those through doi.org so the link is absolute, not relative to /articles/.
  href <- ifelse(is.na(url) | grepl("^https?://", url), url,
          ifelse(grepl("^10\\.[0-9]", url), paste0("https://doi.org/", url),
                 paste0("https://", url)))
  ifelse(is.na(url) | !nzchar(url), "", sprintf("[source](%s)", href))
}
```

```{=html}
<style>
/* This page only: the manifest tables are wide, so widen the layout and let the
   tables use the extra room. The max-width cap is a no-op on narrow screens
   (the container is already smaller than the cap there), so no media query is
   needed. Prose keeps a readable line length. */
/* !important overrides Bootstrap's responsive .container max-width, which is set
   inside a min-width media query and would otherwise cap this page. */
.container.template-article { max-width: 1560px !important; }
.template-article main > p,
.template-article main > div.sourceCode { max-width: 900px; }
.template-article main table { width: 100%; }
</style>
```

taxify resolves names against local taxonomic **backbones** and joins external
**enrichment** datasets that carry **traits**. This page lists all three, straight
from the manifest this version ships. In an R session, the same content is one call
away:

```{r echo = TRUE, eval = FALSE}
taxify_databases()   # backbones + enrichments in one frame
list_backbones()     # the taxonomic backbones
list_enrichments()   # the trait / status datasets
list_traits()        # the cross-source trait vocabulary add_trait() draws on
```

## Taxonomic backbones

Each is a pre-built `.vtr` downloaded once and matched locally. Pass several to
`taxify()` and they form a fallback chain.

```{r}
bb <- list_backbones(verbose = FALSE)
bb$source <- link(bb$source)
knitr::kable(
  bb[, c("name", "scope", "n_names", "size_mb", "version", "source")],
  col.names = c("Backbone", "Scope", "Names", "MB", "Version", ""),
  format.args = list(big.mark = ","),
  align = c("l", "l", "r", "r", "l", "l")
)
```

## Enrichment datasets

Each `add_<source>()` door joins one dataset to a `taxify()` result on the accepted
name. `add_trait()` gathers a single trait across every source that carries it. Browse
a dataset's columns in R with `enrichment_cols("<name>")`.

```{r}
en <- list_enrichments(verbose = FALSE)
en <- en[order(en$name), ]
tc <- ifelse(is.na(en$trait_cols), "", en$trait_cols)
en$n_cols <- vapply(strsplit(tc, ",\\s*"),
                    function(x) sum(nzchar(x)), integer(1))
en$source_url <- link(en$source_url)
knitr::kable(
  en[, c("name", "version", "nrow", "n_cols", "source_url")],
  col.names = c("Enrichment", "Version", "Rows", "Columns", ""),
  format.args = list(big.mark = ","),
  align = c("l", "l", "r", "r", "l"),
  row.names = FALSE
)
```

## Cross-source traits

Name one trait; `add_trait()` finds and harmonizes it from the sources below.

```{r}
tr <- list_traits()
tr <- tr[order(tr$trait), ]
knitr::kable(
  tr[, c("trait", "label", "kind", "unit", "n_sources", "sources")],
  col.names = c("Trait", "Label", "Kind", "Unit", "Sources", "From"),
  align = c("l", "l", "l", "l", "r", "l"),
  row.names = FALSE
)
```
